---
title: "Air Traffic Disruption Thesis"
author: "Anmol Sapru"
date: "2026-02-12"
format:
  pdf:
    toc: true
    number-sections: true
    pdf-engine: xelatex
    margin: 1in
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(airportr)
library(maps)
library(geosphere)
library(httr)
library(jsonlite)
library(fuzzyjoin)
library(rvest)
library(lutz)
library(hms)
library(mgcv)
library(slider)
library(mapproj)
library(knitr)
library(truncnorm)
library(foreach)
library(doParallel)
```

## Why Care?

The country's air transportation is highly interconnected, but thus vulnerable to disruptions. These disruptions lead to delays that expand across the network and affect airlines, airports, and passengers in complex ways. Each day, the United States has 2 million enplanements—which means the impact of even small disturbances can be widespread. Delays can be caused by a plethora of items, such as passengers arriving late, air traffic control holds, and weather.

This project aims to build a simulation that can model such disruptions at scale, thinking from a point of view of robustness, recovery, and systemic impacts—with a final goal of understanding what airport as a hub can help mitigate some of these issues. A goal of the simulation will be replication of a weather event.

Air traffic disruptions have been studied using network theory, agent-based modeling, and operational research. These have shown that disruptions can cascade rapidly. Airlines, airports, and passengers all make interdependent decisions during disruptions, which makes modeling challenging. Existing data sources provide partial insights into these, but using more data and creating a clearer way to simulate is a research gap that can be filled and further used to expand the field of research in aviation.

## Data

Finding the proper data required going down multiple paths, since I knew that no one dataset would be perfect for what I needed. This included examining Global Distribution System data (cost-prohibitive, but would require near-zero cleaning) and data from the Department of Transportation (no data that gave information transformable to departure and arrivals), among others.

As I looked through the trade-offs of the data, I decided to sacrifice some perfection in the scheduling information to get more reproducibility and a lowered cost. Removal of the proprietary data sets would make the results of the projects able to be easily tried over other time periods or region.

Therefore, the work incorporated OpenSky data, which provides access to live and historical flight information. This allowed the simulation to reflect realistic aircraft positioning, which was used to create departure and arrival schedules. By leveraging their Trino database, I was able to get historical flight data, utilising a pull from October 6, 2025 and October 27, 2025. This was a relatively calm period for aviation in the contiguous United States, with few weather events and avoiding the Federal Aviation Administration's Air Traffic Control shutdown, giving me a clean slate.

A balance was made between the hub-and-spoke and point-to-point flying airlines of the United States. The cutoff used were airlines with at least 100 airplanes in their fleets: Alaska Airlines, American Airlines, Delta Air Lines, United Airlines, Frontier Airlines, JetBlue, Southwest Airlines, and Spirit Airlines. Some of these mainline airlines also have regional carriers that fly, which were pulled from OpenSky too.

A key problem that existed with this data was a spike around midnight, where the flights that left before 11 PM EST but did not land before 12 PM EST were erroneous given a departure time of midnight the next day. This shortened the flight time and required some changes to the scheduling for point-to-point flights.

### General Data Cleaning

```{r commercial-airports}
us_airports <- 
  readr::read_csv("https://davidmegginson.github.io/ourairports-data/airports.csv") |>
  filter(
    iso_country == "US",
    scheduled_service == "yes",
    type %in% c("large_airport", "medium_airport"),
    !is.na(ident)
  ) |>
  transmute(
    airport_name = name,
    ICAO = stringr::str_to_upper(stringr::str_trim(ident)),
    IATA = stringr::str_to_upper(stringr::str_trim(iata_code)),
    latitude_deg, longitude_deg
  )
```

After pulling from OpenSky, I needed to do some cleaning of the data. First was making sure the airports were correct. The only airports used were those with ICAO codes that began with 'K', which represents the contiguous United States. Since OpenSky uses the transponder location instead of the schedule data there were some assumed departure and arrival airports that did not make much sense. Some replacement of airports needed to be done and I only used airports from OurAirports that were sized as large or medium with scheduled service—after checking the small airports and determining to not be relevant airports for scheduled service of the airlines.

```{r airport-map}
us_states <- map_data("state")

us_airports |>
  filter(latitude_deg > 24 & latitude_deg < 50, longitude_deg > -125 & longitude_deg < -66) |>
  ggplot() +
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white") +
  geom_point(aes(x = longitude_deg, y = latitude_deg), size = 1.5) +
  coord_map() + 
  theme_void() +
  labs(title = "Scheduled Service Airports (US Lower 48)")

us_airports <- us_airports |>
  select(-latitude_deg, -longitude_deg)
```

```{r load-data, warning = FALSE}
path <- "opensky_daily_exports"
csv_files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)
data_list <- lapply(csv_files, read.csv)
all_data <- do.call(rbind, data_list)
```

```{r remove-twin-aisles}
planes <- 
  read_csv("https://s3.opensky-network.org/data-samples/metadata/aircraft-database-complete-2025-02.csv") |>
  filter(`'icao24'` %in% all_data$icao24)

twin_aisle_check <- all_data |>
  filter(!icao24 %in% planes$`'icao24'`)

twin_aisles <- planes |>
  filter(str_detect(`'model'`, str_c(c("767", "777", "787", "A330", "A350", "310R", "A310"), collapse = "|")))

MAX_RANGE_NM <- 3500

twin_aisle_flights <- twin_aisle_check |>
  left_join(
    airportr::airports |>
      dplyr::select(ICAO, Latitude, Longitude) |>
      rename(lat = Latitude, lon = Longitude) |>
      rename(dep_latitude = lat, dep_longitude = lon), 
    by = c("departure" = "ICAO")
  ) |>
  left_join(
    airportr::airports |>
      dplyr::select(ICAO, Latitude, Longitude) |>
      rename(lat = Latitude, lon = Longitude) |>
      rename(arr_latitude = lat, arr_longitude = lon), 
    by = c("arrival" = "ICAO")
  ) |>
  mutate(
    distance_m = geosphere::distHaversine(
      cbind(dep_longitude, dep_latitude), 
      cbind(arr_longitude, arr_latitude)), 
    distance_nm = distance_m / 1852
  ) |>
  group_by(icao24) |>
  summarize(over_range = any(distance_m > MAX_RANGE_NM * 1852, na.rm = TRUE), .groups = "drop") |>
  filter(over_range) |>
  pull(icao24)
```

I removed twin aisle jets from the data of airplanes, seeing as most of them do travel outside the contiguous United States and the analysis was focused on this area. OpenSky supplies a Complete Aircraft Database that allowed for checking the hexadecimal codes for airplanes to their ownership and type. For other planes not in this data set, a check of if they completed a flight during the period which was over 3500 nautical miles (the current longest single aisle jet flight operated by US airlines—EWR to ARN by United Airlines is just under this threshold).

```{r twin-aisle-visualization}
all_twin_ids <- unique(c(twin_aisles$`'icao24'`, twin_aisle_flights))

map_data_twin <- all_data |>
  filter(icao24 %in% all_twin_ids) |>
  left_join(
    airportr::airports |> select(ICAO, dep_lat = Latitude, dep_lon = Longitude),
    by = c("departure" = "ICAO")
  ) |>
  left_join(
    airportr::airports |> select(ICAO, arr_lat = Latitude, arr_lon = Longitude),
    by = c("arrival" = "ICAO")
  ) |>
  filter(
    !is.na(dep_lat), !is.na(dep_lon), !is.na(arr_lat), !is.na(arr_lon),
    !(dep_lat == arr_lat & dep_lon == arr_lon) 
  )

world_map <- map_data("world")

ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white", linewidth = 0.2) +
  geom_curve(data = map_data_twin, 
             aes(x = dep_lon, y = dep_lat, xend = arr_lon, yend = arr_lat),
             alpha = 0.2, linewidth = 0.2, curvature = 0.2) +
  coord_fixed(1.3) +
  theme_void() +
  labs(
    title = "Routes Flown by Twin Aisle",
    subtitle = paste("Total flights:", nrow(map_data_twin))
  )

# Domestic twin-aisle flights
map_data_domestic <- map_data_twin |>
  filter(
    dep_lon > -126, dep_lon < -66, dep_lat > 24, dep_lat < 50,
    arr_lon > -126, arr_lon < -66, arr_lat > 24, arr_lat < 50
  )

ggplot() +
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group), 
               fill = "gray95", color = "white", linewidth = 0.2) +
  geom_curve(data = map_data_domestic, 
             aes(x = dep_lon, y = dep_lat, xend = arr_lon, yend = arr_lat),
             alpha = 0.3, linewidth = 0.4, curvature = 0.2) +
  coord_fixed(1.3) +
  theme_void() +
  labs(
    title = "Domestic Twin Aisle Flights",
    subtitle = paste("Total flights:", nrow(map_data_domestic))
  )

all_data <- all_data |>
  arrange(icao24) |>
  filter(!icao24 %in% twin_aisle_flights & !icao24 %in% twin_aisles$`'icao24'`)
```

```{r replacing-non-commercials, message = FALSE}
airport_lookup_1 <- read_csv("https://davidmegginson.github.io/ourairports-data/airports.csv") |>
  dplyr::select("ICAO" = "ident", "latitude" = "latitude_deg", "longitude" = "longitude_deg")

download.file("https://nfdc.faa.gov/webContent/28DaySub/extra/22_Jan_2026_APT_CSV.zip", 
              destfile = "APT_CSV.zip", mode = "wb")

contents <- unzip("APT_CSV.zip", list = TRUE)
unzip("APT_CSV.zip", 
      files = contents[grepl("APT_BASE\\.csv$", contents$Name, ignore.case = TRUE), , drop = FALSE]$Name, 
      exdir = "apt_extract")

airport_lookup_2 <- read_csv(
  list.files("apt_extract", pattern = "APT_BASE\\.csv$", recursive = TRUE, 
             full.names = TRUE, ignore.case = TRUE)[1]
) |>
  dplyr::select("ICAO" = "ARPT_ID", "latitude" = "LAT_DECIMAL", "longitude" = "LONG_DECIMAL")

airport_lookups <- bind_rows(airport_lookup_1, airport_lookup_2) |>
  distinct(ICAO, .keep_all = TRUE)

commercial_coords <- airportr::airports |>
  dplyr::select(ICAO, lat = Latitude, lon = Longitude) |>
  mutate(ICAO = str_to_upper(str_trim(ICAO))) |>
  distinct() |>
  semi_join(tibble(ICAO = str_to_upper(str_trim(us_airports$ICAO))), by = "ICAO") |>
  filter(!is.na(lat), !is.na(lon))

lookup_coords <- airport_lookups |>
  transmute(from_icao = str_to_upper(str_trim(ICAO)), 
            lat = as.numeric(latitude), lon = as.numeric(longitude)) |>
  distinct(from_icao, .keep_all = TRUE) |>
  filter(!is.na(from_icao), !is.na(lat), !is.na(lon))

nearest_commercial <- function(lat, lon, comm_tbl, limit_meters = 40000) {
  d_m <- geosphere::distHaversine(cbind(comm_tbl$lon, comm_tbl$lat), c(lon, lat))
  min_dist <- min(d_m)
  i <- which.min(d_m)
  
  if (min_dist <= limit_meters) {
    tibble(nearest_icao = comm_tbl$ICAO[i], dist_m = min_dist)
  } else {
    tibble(nearest_icao = NA_character_, dist_m = min_dist)
  }
}

closest_map_messy <- lookup_coords |>
  mutate(nn = purrr::pmap(list(lat, lon), \(lat, lon)
    nearest_commercial(lat, lon, commercial_coords, limit_meters = 40000)
  )) |>
  unnest(nn) |>
  filter(!from_icao %in% commercial_coords$ICAO, !is.na(nearest_icao))
```

The replacement was done with a simple closest airport calculation for the non-OurAirports ICAOs. In addition to using the OurAirports, the Federal Aviation Administration's Airport Master Record has location about some of the airfields. The longitude and latitudes were calculated to those in OurAirports, and then setting a limit to within 40 kilometers—with anything further being removed due to either being too uncertain or potentially an international flight that was mistakenly categorized as flying domestically. The longitude and latitude for the commercial airports was gotten through the airportr package.

```{r catchment-visualization}
target_hubs <- c("KBDL", "KHVN", "KORH", "KHYA", "KEWB", "KBOS", "KBED", 
                 "KACK", "KPVD", "KWST", "KLEB", "KMHT", "KPSM", "KBTV", 
                 "KRUT", "KAUG", "KBGR", "KBHB", "KPQI", "KPWM", "KRKD")

plot_data <- closest_map_messy |>
  filter(nearest_icao %in% target_hubs) |>
  rename(from_lat = lat, from_lon = lon) |>
  left_join(commercial_coords, by = c("nearest_icao" = "ICAO")) |>
  rename(to_lat = lat, to_lon = lon)

ne_state_names <- c("massachusetts", "connecticut", "rhode island", 
                    "new hampshire", "vermont", "maine")
ne_map_bg <- map_data("state", region = ne_state_names)

ggplot() +
  geom_polygon(data = ne_map_bg, aes(x = long, y = lat, group = group), 
               fill = "gray95", color = "gray60") +
  geom_segment(data = plot_data, 
               aes(x = from_lon, y = from_lat, xend = to_lon, yend = to_lat),
               color = "steelblue", alpha = 0.4, linewidth = 0.3) +
  geom_point(data = plot_data, aes(x = from_lon, y = from_lat), 
             color = "steelblue", size = 0.5, alpha = 0.6) +
  geom_point(data = plot_data |> distinct(nearest_icao, to_lon, to_lat), 
             aes(x = to_lon, y = to_lat), color = "black", size = 2) +
  coord_map() +
  theme_void() +
  labs(title = "Airport Catchment Areas", subtitle = "Mappings to New England Airports")
```

```{r cleaning-airlines-airports}
all_data <- all_data |>
  left_join(closest_map_messy, by = c("departure" = "from_icao")) |>
  mutate(departure = if_else(!is.na(nearest_icao), nearest_icao, departure)) |>
  dplyr::select(-nearest_icao, -dist_m) |>
  left_join(closest_map_messy, by = c("arrival" = "from_icao")) |>
  mutate(arrival = if_else(!is.na(nearest_icao), nearest_icao, arrival)) |>
  dplyr::select(-nearest_icao, -dist_m)

flights <- all_data |>
  filter(
    departure %in% us_airports$ICAO, 
    arrival %in% us_airports$ICAO, 
    departure != arrival,
    str_sub(callsign, 1, 3) %in% c("AAL", "ASA", "DAL", "EDV", "FFT", "JBU", "NKS", "RPA", 
                                   "SKW", "SWA", "UAL", "QXE", "JIA", "ENY", "PDT")
  ) |>
  mutate(
    firstseen = ymd_hms(firstseen),
    lastseen = ymd_hms(lastseen),
    duration = as.numeric(difftime(lastseen, firstseen, units = "secs")) + 30 * 60,
    lastseen = firstseen + duration,
    route_from = pmin(departure, arrival),
    route_to = pmax(departure, arrival)
  )
```

```{r add-timezones, warning = FALSE}
airports_ref <- airportr::airports |>
  dplyr::select(ICAO, lat = Latitude, lon = Longitude) |>
  distinct()

tz_lookup <- airports_ref |>
  rowwise() |>
  mutate(timezone = tz_lookup_coords(lat, lon, method = "fast")) |>
  ungroup()

flights <- flights |>
  mutate(
    firstseen = lubridate::ymd_hms(firstseen, tz = "UTC"),
    lastseen = lubridate::ymd_hms(lastseen, tz = "UTC")
  )

dep_tz_tbl <- tz_lookup |> dplyr::select(ICAO, departure_tz = timezone)
arr_tz_tbl <- tz_lookup |> dplyr::select(ICAO, arrival_tz = timezone)

flights <- flights |>
  left_join(dep_tz_tbl, by = c("departure" = "ICAO")) |>
  left_join(arr_tz_tbl, by = c("arrival" = "ICAO")) |>
  mutate(
    firstseen = as.POSIXct(firstseen, tz = "UTC"),
    lastseen = as.POSIXct(lastseen, tz = "UTC"),
    departure_tz = departure_tz |> as.character() |> stringr::str_trim(),
    arrival_tz = arrival_tz |> as.character() |> stringr::str_trim(),
    departure_tz = dplyr::if_else(is.na(departure_tz) | !departure_tz %in% OlsonNames(), "UTC", departure_tz),
    arrival_tz = dplyr::if_else(is.na(arrival_tz) | !arrival_tz %in% OlsonNames(), "UTC", arrival_tz)
  ) |>
  group_by(departure_tz) |>
  mutate(dep_local = lubridate::with_tz(firstseen, unique(departure_tz))) |>
  ungroup() |>
  group_by(arrival_tz) |>
  mutate(arr_local = lubridate::with_tz(lastseen, unique(arrival_tz))) |>
  ungroup() |>
  mutate(
    dep_minute = lubridate::hour(dep_local) * 60 + lubridate::minute(dep_local),
    arr_minute = lubridate::hour(arr_local) * 60 + lubridate::minute(arr_local),
    dep_hour = lubridate::hour(dep_local),
    arr_hour = lubridate::hour(arr_local)
  )
```

Other data cleaning involved lots of work with the departure and arrival timezones. First, a reference table of the airports from the airportr package was built, which included timezone was determined by passing its latitude and longitude through the lutz package. With valid timezones in place, the UTC-stamped firstseen and lastseen columns could be converted to local departure and arrival times.

```{r turnaround-distribution}
turnaround_times <- flights |>
  arrange(icao24, firstseen) |>
  group_by(icao24) |>
  mutate(turnaround_mins = (as.numeric(firstseen) - lag(as.numeric(lastseen))) / 60) |>
  ungroup() |>
  filter(!is.na(turnaround_mins), turnaround_mins > 0, turnaround_mins < 120) |>
  pull(turnaround_mins)

fit <- MASS::fitdistr(turnaround_times, densfun = "gamma")
shape_hat <- unname(fit$estimate["shape"])
rate_hat <- unname(fit$estimate["rate"])

cat("Gamma shape:", shape_hat, "\nGamma rate:", rate_hat, "\n")
```

The turnaround time of an airplane was modeled as a Gamma distribution, which makes sense as its definition of "waiting time". For each aircraft, identified by its unique icao24 transponder code, flights were sorted chronologically and the gap between consecutive flights was calculated as the difference between one departure and its previous arrival. These gaps were filtered to exclude negative values and those exceeding two hours, focusing on ground turnaround times instead of overnight stays or maintenance stops.

```{r turnaround-hist-fit, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
xmax <- quantile(turnaround_times, 0.99, na.rm = TRUE)

ggplot(tibble(turnaround_mins = turnaround_times), aes(x = turnaround_mins)) +
  geom_histogram(aes(y = after_stat(density)), bins = 35, boundary = 0, closed = "left") +
  stat_function(
    fun = \(x) dgamma(x, shape = shape_hat, rate = rate_hat),
    linewidth = 1
  ) +
  coord_cartesian(xlim = c(0, xmax)) +
  labs(
    title = "Turnaround time distribution with fitted Gamma",
    x = "Turnaround time (minutes)",
    y = "Density"
  ) +
  theme_minimal()
```

## Hub-and-Spoke Carriers

For hub and spoke airlines, some key ideas are the hubs they have, the banks they operate at airports, and destinations flown to.

```{r select-hubs}
target_airlines <- c("UAL", "DAL", "AAL", "ASA")

hub_map <- flights |>
  mutate(airline = stringr::str_sub(callsign, 1, 3)) |>
  filter(airline %in% target_airlines) |>
  count(airline, hub = departure, name = "flight_volume") |>
  group_by(airline) |>
  slice_max(order_by = flight_volume, n = 12, with_ties = FALSE) |>
  ungroup()

own_hubs_tbl <- hub_map |>
  distinct(airline, dest = hub) |>
  mutate(is_own_hub = TRUE)

hub_map |>
  arrange(airline, desc(flight_volume)) |>
  group_by(airline) |>
  mutate(rank = row_number()) |>
  ungroup() |>
  select(airline, rank, hub) |>
  tidyr::pivot_wider(names_from = airline, values_from = hub) |>
  arrange(rank) |>
  select(-rank) |>
  knitr::kable(
    col.names = c("United (UAL)", "Delta (DAL)", "American (AAL)", "Alaska (ASA)"),
    caption = "Top 12 hubs by Departure Volume"
  )
```

The top 12 airports for each airlines were selected as hubs. For airlines (such as United) that did not have this many airports, the future results of the code were not affected by this overestimate, since flights between hubs were scheduled in the same way as those from a hub to spoke.

```{r add-regionals-to-mainlines, warning = FALSE}
regional_flights <- flights |>
  filter(str_detect(callsign, "^(RPA|SKW)")) |>
  pivot_longer(c(departure, arrival), names_to = "leg", values_to = "airport") |>
  inner_join(hub_map, by = c("airport" = "hub")) |>
  count(icao24, airline, name = "hub_hits") |>
  group_by(icao24) |>
  slice_max(hub_hits, n = 1, with_ties = FALSE) |>
  ungroup()

flights <- flights |>
  left_join(regional_flights, by = join_by(icao24)) |>
  mutate(
    callsign = case_when(
      !is.na(airline) ~ str_c(airline, str_sub(callsign, 4)),
      str_sub(callsign, 1, 3) == "EDV" ~ str_c("DAL", str_sub(callsign, 4)),
      str_sub(callsign, 1, 3) == "QXE" ~ str_c("ASA", str_sub(callsign, 4)),
      str_sub(callsign, 1, 3) %in% c("JIA", "ENY", "PDT") ~ str_c("AAL", str_sub(callsign, 4)),
      str_sub(callsign, 1, 3) %in% c("ASH", "UCA", "GJS") ~ str_c("UAL", str_sub(callsign, 4)),
      TRUE ~ callsign
    ),
    airline = str_sub(callsign, 1, 3)
  ) |>
  dplyr::select(-hub_hits)
```

The regional airlines are a huge part of the networks that are being flown. Some of the regional airlines only fly for a single larger counterpart (Endevour for Delta or Horizon for Alaska), but the larger Republic and Skywest fly for multiple airlines. Therefore, the number of hubs that each airplane takes-off or lands at was used to determine what marketable airline each one belonged to.

```{r helper-functions}
# Operating hours
op_start <- 5
op_end <- 23

# Bank detection from GAM-smoothed curve
banks_from_curve_topk_hours <- function(curve_df, op_start = 5, op_end = 23, win = 1, 
                                        min_banks = 3, max_banks = 4) {
  x <- curve_df |>
    arrange(hour) |>
    mutate(
      lam_s = slider::slide_dbl(lambda, mean, .before = win, .after = win, .complete = TRUE),
      prev = lag(lam_s),
      nextv = lead(lam_s),
      valley = !is.na(prev) & !is.na(nextv) & lam_s < prev & lam_s < nextv,
      depth = ((prev + nextv) / 2) - lam_s
    )

  valleys <- x |> filter(valley) |> arrange(desc(depth))

  n_cuts <- case_when(
    nrow(valleys) >= (max_banks - 1) ~ (max_banks - 1),
    nrow(valleys) >= (min_banks - 1) ~ (min_banks - 1),
    TRUE ~ NA_integer_
  )

  if (is.na(n_cuts)) {
    cuts <- seq(op_start, op_end, length.out = min_banks + 1)
  } else {
    cuts_mid <- valleys |> slice_head(n = n_cuts) |> pull(hour)
    cuts <- sort(unique(c(op_start, cuts_mid, op_end)))
    if (length(cuts) < (min_banks + 1)) {
      cuts <- seq(op_start, op_end, length.out = min_banks + 1)
    }
  }

  tibble::tibble(
    start_hour = cuts[-length(cuts)],
    end_hour = cuts[-1],
    bank_idx = seq_along(cuts[-1]),
    bank = paste0("Bank ", bank_idx),
    start_min = start_hour * 60,
    end_min = end_hour * 60,
    n_banks = length(cuts) - 1
  )
}

# Fit GAM curve
fit_curve <- function(count_df) {
  fit <- mgcv::gam(
    y ~ s(hour, bs = "cs", k = 10),
    family = poisson(),
    data = count_df,
    method = "REML"
  )
  tibble(hour = 0:23, lambda = predict(fit, newdata = tibble(hour = 0:23), type = "response"))
}

# Connectivity score for slot optimization
connect_score <- function(t, arr_vec, mct = 35, max_ct = 180) {
  t <- as.integer(t) %% 1440
  lo <- t - max_ct
  hi <- t - mct
  rng <- seq(lo, hi)
  rng_mod <- (rng %% 1440) + 1
  sum(arr_vec[rng_mod])
}

# Optimize departure slots within a bank
optimize_slots_in_bank <- function(start_min, end_min, n_slots, arr_vec,
                                   mct = 35, max_ct = 180, buffer = 5, 
                                   min_sep = 4, search_radius = 30, passes = 2) {
  if (n_slots <= 0) return(integer())
  lo <- start_min + buffer
  hi <- end_min - buffer
  if (hi <= lo) return(rep(round((start_min + end_min) / 2), n_slots))

  slots <- if (n_slots == 1) {
    round((lo + hi) / 2)
  } else {
    round(seq(lo, hi, length.out = n_slots))
  }

  for (p in seq_len(passes)) {
    for (i in seq_along(slots)) {
      left_bound <- if (i == 1) lo else slots[i - 1] + min_sep
      right_bound <- if (i == length(slots)) hi else slots[i + 1] - min_sep
      if (right_bound < left_bound) next

      cand_lo <- max(left_bound, slots[i] - search_radius)
      cand_hi <- min(right_bound, slots[i] + search_radius)
      cands <- cand_lo:cand_hi

      scores <- vapply(cands, \(tt) connect_score(tt, arr_vec, mct = mct, max_ct = max_ct), numeric(1))
      slots[i] <- cands[which.max(scores)]
    }
  }
  slots
}

# Allocate destinations within a bank
allocate_dests_bank <- function(n_slots, dests, probs) {
  if (n_slots <= 0) return(character())
  if (is.null(dests) || length(dests) == 0) return(rep(NA_character_, n_slots))

  probs <- as.numeric(probs)
  probs[is.na(probs) | probs < 0] <- 0
  if (sum(probs) <= 0) return(rep(NA_character_, n_slots))
  probs <- probs / sum(probs)

  exp_ct <- n_slots * probs
  base <- floor(exp_ct)
  rem <- n_slots - sum(base)
  frac <- exp_ct - base

  if (rem > 0) {
    add_idx <- order(frac, decreasing = TRUE)[seq_len(rem)]
    base[add_idx] <- base[add_idx] + 1L
  }

  vec <- rep(dests, base)
  sample(vec, size = n_slots, replace = FALSE)
}

# Get duration safely
get_duration_safe <- function(al, dep, arr, r_block) {
  match_row <- r_block |> filter(airline == al, departure == dep, arrival == arr)
  if (nrow(match_row) > 0) return(as.integer(match_row$dur_sec_med))
  return(as.integer(3600))
}

# Add coordinates to schedule
add_coords_11cols <- function(tbl, airports_ref) {
  dep_coords <- airports_ref |> 
    dplyr::select(ICAO, dep_latitude = lat, dep_longitude = lon) |> distinct()
  arr_coords <- airports_ref |> 
    dplyr::select(ICAO, arr_latitude = lat, arr_longitude = lon) |> distinct()
  
  tbl |> 
    left_join(dep_coords, by = c("departure" = "ICAO")) |> 
    left_join(arr_coords, by = c("arrival" = "ICAO")) |> 
    dplyr::select(
      any_of(c("airline", "callsign", "tail_number", "departure", "arrival", 
               "day_of_week", "departure_time", "arrival_time")),
      dep_latitude, dep_longitude, arr_latitude, arr_longitude
    )
}
```

To understand the banking at different hubs, a GAM with a cyclic cubic spline is fit to hourly flight counts, which gives a smooth Poisson rate curve across the day. This curve is then passed to an algorithm that identifies the peaks and throughs The deepest valleys serve as cut points that partition the operating day (defined as 5:00 to 23:00 local time) into departure banks, allowing to mimick the wave structure characteristic of hub-and-spoke scheduling.

Within each bank, departure slots are optimized using a connectivity scoring function. For a candidate departure time, the score counts how many arriving flights fall within a feasible connection window—after the minimum connect time of 35 minutes but before the maximum of 180 minutes—giving preference to departure times that maximize the number of passengers who can connect from inbound flights. The optimizer initializes slots at evenly spaced positions within each bank and then iteratively shifts each slot within a local search radius to maximize its connectivity score, respecting minimum separation constraints between adjacent slots.

Destinations are then allocated to slots within each bank using a proportional assignment method. Each destination receives a number of slots proportional to its observed frequency, with fractional remainders resolved by a largest-remainder rounding scheme. The resulting assignments are shuffled randomly to avoid systematic ordering effects. Finally, utility functions handle flight duration lookup from historical medians and coordinate enrichment by joining departure and arrival airport positions for downstream mapping.

```{r prepare-hub-spoke-data}
wave_carriers <- c("DAL", "AAL", "ASA", "UAL")

valid_hubs <- hub_map |>
  filter(airline %in% wave_carriers) |>
  distinct(airline, hub)

flights_wave <- flights |>
  filter(airline %in% wave_carriers) |>
  mutate(
    dep_local = lubridate::with_tz(firstseen, departure_tz),
    arr_local = lubridate::with_tz(lastseen, arrival_tz),
    dep_minute = lubridate::hour(dep_local) * 60 + lubridate::minute(dep_local),
    arr_minute = lubridate::hour(arr_local) * 60 + lubridate::minute(arr_local),
    dep_hour = lubridate::hour(dep_local),
    arr_hour = lubridate::hour(arr_local),
    dep_date = as.Date(dep_local),
    arr_date = as.Date(arr_local)
  )

# Fit curves for departures and arrivals
dep_counts <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  count(airline, hub = departure, hour = dep_hour, name = "y") |>
  complete(nesting(airline, hub), hour = 0:23, fill = list(y = 0))

arr_counts <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "arrival" = "hub")) |>
  count(airline, hub = arrival, hour = arr_hour, name = "y") |>
  complete(nesting(airline, hub), hour = 0:23, fill = list(y = 0))

dep_curves <- dep_counts |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y) fit_curve(.x)) |>
  ungroup()

dep_banks <- dep_counts |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y) {
    curve <- fit_curve(.x)
    banks_from_curve_topk_hours(curve, win = 1, min_banks = 3, max_banks = 4)
  }) |>
  ungroup()

# Route block times
route_block <- flights_wave |>
  mutate(dur_sec = as.numeric(difftime(lastseen, firstseen, units = "secs"))) |>
  filter(is.finite(dur_sec), dur_sec > 10 * 60, dur_sec < 10 * 3600) |>
  group_by(airline, departure, arrival) |>
  summarise(n = n(), dur_sec_med = median(dur_sec, na.rm = TRUE), .groups = "drop")
```

```{r hub-spoke-visualization}
top_hubs <- dep_counts |>
  group_by(hub) |>
  summarise(total = sum(y), .groups = "drop") |>
  slice_max(total, n = 12) |>
  pull(hub)

rects <- dep_banks |> filter(hub %in% top_hubs) |>
  mutate(xmin = start_hour, xmax = end_hour, ymin = -Inf, ymax = Inf)

ggplot() +
  geom_rect(data = rects, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, group = bank_idx),
            alpha = 0.10, inherit.aes = FALSE) +
  geom_col(data = dep_counts |> filter(hub %in% top_hubs),
           aes(x = hour, y = y, group = airline), alpha = 0.30, width = 0.9, position = "identity") +
  geom_line(data = dep_curves |> filter(hub %in% top_hubs),
            aes(x = hour, y = lambda, group = airline), linewidth = 0.9) +
  scale_x_continuous(breaks = 0:23) +
  facet_wrap(~ hub, scales = "free_y") +
  labs(title = "Departures by airport: waves + banks", x = "Local hour", 
       y = "Observed departures (bars) / Estimated intensity λ (line)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "none")
```

```{r generate-7-day-hub-spoke-schedule}
# Daily demand by weekday
hub_dep_obs_daily <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  left_join(dep_banks |> dplyr::select(airline, hub, bank, bank_idx, start_min, end_min),
            by = c("airline", "departure" = "hub")) |>
  filter(!is.na(start_min), !is.na(end_min), dep_minute >= start_min, dep_minute < end_min) |>
  mutate(hub = departure, day_name = lubridate::wday(dep_date, label = TRUE, abbr = FALSE)) |>
  group_by(airline, hub, bank, bank_idx, start_min, end_min, day_name, dep_date) |>
  summarise(n_day = n(), .groups = "drop") |>
  group_by(airline, hub, bank, bank_idx, start_min, end_min, day_name) |>
  summarise(n_dep = as.integer(ceiling(stats::quantile(n_day, 0.90, na.rm = TRUE))), .groups = "drop") |>
  mutate(n_dep = pmax(0L, n_dep))

# Route probabilities
route_shares <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  left_join(dep_banks |> dplyr::select(airline, hub, bank, bank_idx, start_min, end_min),
            by = c("airline", "departure" = "hub")) |>
  filter(dep_minute >= start_min, dep_minute < end_min) |>
  mutate(hub = departure, dest = arrival) |>
  count(airline, hub, bank, bank_idx, dest, name = "n") |>
  left_join(own_hubs_tbl, by = c("airline", "dest")) |>
  mutate(is_own_hub = coalesce(is_own_hub, FALSE), w = if_else(is_own_hub, 0.5, 1.0), n_w = n * w) |>
  group_by(airline, hub, bank, bank_idx) |>
  mutate(p = n_w / sum(n_w)) |>
  ungroup()

route_dist <- route_shares |>
  group_by(airline, hub, bank, bank_idx) |>
  summarise(dests = list(dest), probs = list(p), .groups = "drop")

# Arrival profiles
arr_vec_tbl <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "arrival" = "hub")) |>
  mutate(hub = arrival) |>
  group_by(airline, hub) |>
  summarise(
    n_days = n_distinct(arr_date),
    arr_vec = list(tabulate((arr_minute %% 1440) + 1L, nbins = 1440) / pmax(1L, n_days)),
    .groups = "drop"
  )

# Daily schedule generator
generate_daily_schedule <- function(target_day, obs_data, routes_data, arr_vec_data) {
  day_obs <- obs_data |> filter(day_name == target_day)
  if (nrow(day_obs) == 0) return(tibble())

  day_schedule <- day_obs |>
    left_join(arr_vec_data, by = c("airline", "hub")) |>
    mutate(
      slot_minute = purrr::pmap(
        list(start_min, end_min, n_dep, arr_vec),
        \(.s, .e, .n, .v) optimize_slots_in_bank(.s, .e, .n, .v, mct = 35, max_ct = 180,
                                                   buffer = 5, min_sep = 4, search_radius = 30, passes = 2)
      )
    ) |>
    tidyr::unnest(slot_minute) |>
    mutate(slot_minute = as.integer(round(slot_minute))) |>
    group_by(airline, hub, bank, bank_idx) |>
    mutate(slot_idx = row_number()) |>
    ungroup() |>
    left_join(routes_data, by = c("airline", "hub", "bank", "bank_idx")) |>
    group_by(airline, hub, bank, bank_idx) |>
    mutate(
      dest_vec = list(allocate_dests_bank(first(n_dep), dests[[1]], probs[[1]])),
      dest = dest_vec[[1]][slot_idx]
    ) |>
    ungroup() |>
    dplyr::select(-dests, -probs, -dest_vec) |>
    filter(!is.na(dest))

  base_demand <- day_schedule |>
    mutate(dep_sec = as.integer(round(slot_minute)) * 60L) |>
    rowwise() |>
    mutate(duration_out = get_duration_safe(airline, hub, dest, route_block)) |>
    ungroup() |>
    transmute(
      airline, flight_id_base = paste0(airline, "-", hub, "-", bank_idx, "-", row_number()),
      origin = hub, dest = dest, dep_time = dep_sec, arr_time = dep_sec + duration_out
    )

  return_demand <- base_demand |>
    mutate(origin_ret = dest, dest_ret = origin, dep_time_ret = arr_time + 45 * 60) |>
    rowwise() |>
    mutate(duration_in = get_duration_safe(airline, origin_ret, dest_ret, route_block)) |>
    ungroup() |>
    transmute(
      airline, flight_id_base = paste0(flight_id_base, "R"),
      origin = origin_ret, dest = dest_ret,
      dep_time = dep_time_ret, arr_time = dep_time_ret + duration_in
    )

  bind_rows(base_demand, return_demand) |> mutate(day_of_week = target_day) |> arrange(dep_time)
}

# Run 7-day simulation
set.seed(1)

fleet_state <- tibble(airline = character(), tail_num = character(), loc = character(), ready_at = integer())
tail_counter <- 1
turn_buffer_hub <- 45 * 60
seconds_per_day <- 24 * 60 * 60

days_to_sim <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
full_week_results <- list()

cat("Starting 7-day hub-spoke simulation...\n")

for (d_idx in seq_along(days_to_sim)) {
  current_day <- days_to_sim[d_idx]
  cat("Simulating:", current_day, "\n")
  
  daily_flights <- generate_daily_schedule(current_day, hub_dep_obs_daily, route_dist, arr_vec_tbl)
  
  if (nrow(fleet_state) > 0) {
    fleet_state <- fleet_state |> mutate(ready_at = pmax(0, ready_at - seconds_per_day))
  }
  
  assigned_day <- list()
  daily_flights <- daily_flights |> arrange(dep_time)
  
  if (nrow(daily_flights) > 0) {
    for (i in seq_len(nrow(daily_flights))) {
      flight <- daily_flights[i, ]
      
      candidates <- fleet_state |>
        filter(airline == flight$airline, loc == flight$origin, ready_at <= flight$dep_time) |>
        arrange(desc(ready_at))
      
      if (nrow(candidates) > 0) {
        chosen_tail <- candidates$tail_num[1]
        fleet_state <- fleet_state |>
          filter(tail_num != chosen_tail) |>
          add_row(airline = flight$airline, tail_num = chosen_tail, 
                  loc = flight$dest, ready_at = flight$arr_time + turn_buffer_hub)
      } else {
        chosen_tail <- paste0("N", stringr::str_pad(tail_counter, 5, pad = "0"), flight$airline)
        tail_counter <- tail_counter + 1
        fleet_state <- fleet_state |>
          add_row(airline = flight$airline, tail_num = chosen_tail, 
                  loc = flight$dest, ready_at = flight$arr_time + turn_buffer_hub)
      }
      
      assigned_day[[i]] <- flight |> mutate(tail_number = chosen_tail)
    }
    full_week_results[[d_idx]] <- bind_rows(assigned_day)
  }
}

final_week_schedule <- bind_rows(full_week_results)

# Format final schedule
final_formatted <- final_week_schedule |>
  rename(departure = origin, arrival = dest) |>
  mutate(
    callsign = paste0(airline, str_sub(tail_number, 2, 5)),
    departure_time = hms::as_hms(dep_time),
    arrival_time = hms::as_hms(arr_time),
    day_of_week = factor(day_of_week, levels = days_to_sim, ordered = TRUE)
  ) |>
  add_coords_11cols(airports_ref = airports_ref) |>
  mutate(duration = as.numeric(arrival_time - departure_time)) |> 
  arrange(day_of_week, departure_time)

saveRDS(final_formatted, "hub_spoke_schedule.rds")
cat("Hub-spoke schedule saved:", nrow(final_formatted), "flights\n")
```

The creation of schedule for hub-and-spoke carriers first estimates demand by calculating how many departures each airline needs per bank at each hub on each day of the week, using the 90th percentile of observed daily counts as the target to size the schedule for a busy-but-not-extreme day. It then computes route probability shares for each destination within each bank, downweighting routes to the airline's own hubs by 50% to avoid over-saturating hub-to-hub connections relative to spoke destinations. Alongside this, it builds a minute-resolution vector of average arriving flights at each hub, which the connectivity optimizer later uses to score candidate departure times.

The daily schedule generator brings these inputs together: for a given day, it optimizes departure slot times within each bank to maximize connectivity with inbound flights, assigns destinations to those slots proportionally, looks up historical block times for each route, and generates matching return flights assuming a 45-minute turnaround at the spoke. This generator is then called within a seven-day fleet simulation loop that assigns tail numbers using a greedy algorithm. For each flight, the simulator checks whether an available aircraft of the right airline is already at the origin airport and reuses it if so, otherwise introducing a new tail number into the fleet. Aircraft availability carries over between days, allowing the simulation to track how many unique aircraft each airline actually needs across the full week.

## Point-to-Point Carriers

For hub and spoke airlines, some key ideas are the routes they operate

```{r point-to-point-setup}
ptp_prefix <- c("FFT", "JBU", "NKS", "SWA")

ptp_flights <- all_data |>
  mutate(
    airline = stringr::str_sub(callsign, 1, 3),
    firstseen_utc = lubridate::ymd_hms(firstseen, tz = "UTC"),
    lastseen_utc = lubridate::ymd_hms(lastseen, tz = "UTC"),
    service_date_utc = as.Date(firstseen_utc)
  ) |>
  filter(
    departure %in% us_airports$ICAO,
    arrival %in% us_airports$ICAO,
    departure != arrival,
    airline %in% ptp_prefix
  ) |>
  mutate(
    duration_s = as.numeric(difftime(lastseen_utc, firstseen_utc, units = "secs")),
    duration_m = duration_s / 60
  ) |>
  filter(is.finite(duration_m), duration_m > 0, duration_m < 24 * 60) |>
  dplyr::select(airline, callsign, icao24, departure, arrival, firstseen_utc, 
                lastseen_utc, service_date_utc, duration_m)

# Spread midnight spike
spread_midnight_spike_utc <- function(flights_tbl, airlines = c("SWA", "FFT", "NKS", "JBU"),
                                      max_back_hours = 2L, expected_method = "p50_1to3", seed = 1) {
  set.seed(seed)

  x <- flights_tbl |>
    mutate(
      airline = as.character(airline),
      firstseen_utc = as.POSIXct(firstseen_utc, tz = "UTC"),
      lastseen_utc = as.POSIXct(lastseen_utc, tz = "UTC"),
      service_date_utc = as.Date(firstseen_utc),
      hour_utc = lubridate::hour(firstseen_utc),
      .row_id = row_number()
    )

  hourly <- x |>
    filter(airline %in% airlines) |>
    count(airline, service_date_utc, hour_utc, name = "n") |>
    tidyr::complete(airline, service_date_utc, hour_utc = 0:23, fill = list(n = 0))

  expected <- hourly |>
    group_by(airline, service_date_utc) |>
    summarise(
      n0 = n[hour_utc == 0], n1 = n[hour_utc == 1], n2 = n[hour_utc == 2], n3 = n[hour_utc == 3],
      expected0 = as.integer(stats::median(c(n1, n2, n3))),
      excess0 = pmax(0L, n0 - expected0),
      .groups = "drop"
    ) |>
    filter(excess0 > 0)

  if (nrow(expected) == 0) return(x |> dplyr::select(-hour_utc, -.row_id))

  cand <- x |>
    filter(airline %in% airlines, hour_utc == 0) |>
    inner_join(expected |> dplyr::select(airline, service_date_utc, excess0),
               by = c("airline", "service_date_utc"))

  if (nrow(cand) == 0) return(x |> dplyr::select(-hour_utc, -.row_id))

  movers <- cand |>
    group_by(airline, service_date_utc) |>
    group_modify(\(.df, .key) {
      n_move <- min(nrow(.df), .df$excess0[[1]])
      if (n_move <= 0) return(.df[0, ])
      dplyr::slice_sample(.df, n = n_move)
    }) |>
    ungroup() |>
    dplyr::select(.row_id)

  if (nrow(movers) == 0) return(x |> dplyr::select(-hour_utc, -.row_id))

  ks <- 1L:max_back_hours
  probs <- rev(seq_len(max_back_hours))
  probs <- probs / sum(probs)

  shift_tbl <- movers |> mutate(k = sample(ks, size = n(), replace = TRUE, prob = probs))

  x |>
    left_join(shift_tbl, by = ".row_id") |>
    mutate(
      firstseen_utc = if_else(!is.na(k), firstseen_utc - lubridate::hours(k), firstseen_utc),
      lastseen_utc = if_else(!is.na(k), lastseen_utc - lubridate::hours(k), lastseen_utc),
      service_date_utc = as.Date(firstseen_utc)
    ) |>
    dplyr::select(-hour_utc, -.row_id, -k)
}

ptp_flights <- spread_midnight_spike_utc(ptp_flights, airlines = ptp_prefix, 
                                         max_back_hours = 2, expected_method = "p50_1to3", seed = 1)
```

The previously discussed midnight spike was artificially dealt with here through comparing the flight count in hour zero against the median of hours one through three for each airline and date. Any excess above that expected baseline is identified, and a random sample of that many flights is selected from the midnight hour and shifted backward by one or two hours (with a bias toward smaller shifts).

```{r point-to-point-simulation}
clean_flights <- all_data |>
  mutate(
    airline = stringr::str_sub(callsign, 1, 3),
    firstseen_utc = lubridate::ymd_hms(firstseen, tz = "UTC"),
    lastseen_utc = lubridate::ymd_hms(lastseen, tz = "UTC"),
    duration_min = as.numeric(difftime(lastseen_utc, firstseen_utc, units = "mins"))
  ) |>
  filter(
    departure %in% us_airports$ICAO,
    arrival %in% us_airports$ICAO,
    departure != arrival,
    airline %in% ptp_prefix,
    duration_min > 20, duration_min < 600
  )

fleet_sizes <- clean_flights |>
  group_by(airline) |>
  summarise(n_planes = n_distinct(icao24)) |>
  tibble::deframe()

ptp_route_block_times <- clean_flights |>
  group_by(airline, departure, arrival) |>
  summarise(block_min = as.integer(median(duration_min)), .groups = "drop")

transition_matrix <- clean_flights |>
  count(airline, departure, arrival, name = "flights") |>
  group_by(airline, departure) |>
  mutate(prob = flights / sum(flights)) |>
  ungroup() |>
  dplyr::select(airline, origin = departure, dest = arrival, prob)

starting_weights <- clean_flights |>
  count(airline, departure, name = "weight") |>
  group_by(airline) |>
  mutate(prob = weight / sum(weight)) |>
  ungroup()

simulate_airline <- function(airline_code, n_planes, days = 7) {
  my_trans <- transition_matrix |> filter(airline == airline_code)
  my_blocks <- ptp_route_block_times |> filter(airline == airline_code)
  my_starts <- starting_weights |> filter(airline == airline_code)
  
  if (nrow(my_starts) == 0) return(NULL)
  
  start_sim <- as.POSIXct("2026-01-01 05:00:00", tz = "UTC")
  end_sim <- start_sim + lubridate::days(days)
  
  fleet <- tibble(
    tail_number = sprintf("N%04d%s", 1:n_planes, airline_code),
    airline = airline_code,
    current_airport = sample(my_starts$departure, n_planes, replace = TRUE, prob = my_starts$prob),
    next_available = start_sim + lubridate::minutes(sample(0:240, n_planes, replace = TRUE))
  )
  
  schedule_log <- list()
  cat(sprintf("Simulating %s with %d planes...\n", airline_code, n_planes))
  
  for (p in 1:n_planes) {
    curr_tail <- fleet$tail_number[p]
    curr_loc <- fleet$current_airport[p]
    curr_time <- fleet$next_available[p]
    plane_legs <- list()
    
    while (curr_time < end_sim) {
      routes <- my_trans |> filter(origin == curr_loc)
      
      if (nrow(routes) == 0) {
        curr_loc <- sample(my_starts$departure, 1, prob = my_starts$prob)
        curr_time <- curr_time + lubridate::hours(4)
        next
      }
      
      next_dest <- sample(routes$dest, 1, prob = routes$prob)
      
      blk <- my_blocks |> filter(departure == curr_loc, arrival == next_dest) |> pull(block_min)
      if (length(blk) == 0) blk <- 120
      
      dep_time <- curr_time
      arr_time <- dep_time + lubridate::minutes(blk)
      
      utc_hour <- lubridate::hour(dep_time)
      if (utc_hour >= 6 && utc_hour < 10) {
        dep_time <- dep_time + lubridate::hours(6)
        arr_time <- dep_time + lubridate::minutes(blk)
      }
      
      plane_legs[[length(plane_legs) + 1]] <- tibble(
        airline = airline_code, tail_number = curr_tail,
        departure = curr_loc, arrival = next_dest,
        dep_time = dep_time, arr_time = arr_time, duration = blk
      )
      
      curr_loc <- next_dest
      curr_time <- arr_time + lubridate::minutes(45)
    }
    schedule_log[[p]] <- bind_rows(plane_legs)
  }
  
  bind_rows(schedule_log)
}

sim_results <- list()
for (al in names(fleet_sizes)) {
  sim_results[[al]] <- simulate_airline(al, fleet_sizes[[al]])
}

full_gen_schedule <- bind_rows(sim_results)

airport_locs_clean <- airport_lookup_1 |> 
  transmute(ICAO = ICAO, lat = latitude, lon = longitude) |> 
  filter(!is.na(lat), !is.na(lon)) |> 
  distinct(ICAO, .keep_all = TRUE)

point_to_point_schedule <- full_gen_schedule |>
  mutate(
    dep_time_et = lubridate::with_tz(dep_time, "America/New_York"),
    arr_time_et = lubridate::with_tz(arr_time, "America/New_York")
  ) |>
  filter(dep_time_et < as.POSIXct("2026-01-08", tz = "America/New_York")) |>
  mutate(
    day_of_week = factor(weekdays(dep_time_et), 
                         levels = c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                    "Friday", "Saturday", "Sunday"), ordered = TRUE),
    callsign = paste0(airline, str_sub(tail_number, 2, 5)),
    departure_time = hms::as_hms(dep_time_et),
    arrival_time = hms::as_hms(arr_time_et),
    duration = as.numeric(duration * 60)
  ) |>
  left_join(airport_locs_clean |> rename(dep_latitude = lat, dep_longitude = lon), 
            by = c("departure" = "ICAO")) |>
  left_join(airport_locs_clean |> rename(arr_latitude = lat, arr_longitude = lon), 
            by = c("arrival" = "ICAO")) |>
  dplyr::select(
    airline, callsign, tail_number, departure, arrival,
    day_of_week, departure_time, arrival_time, duration,
    dep_latitude, dep_longitude, arr_latitude, arr_longitude
  ) |>
  arrange(day_of_week, departure_time)

saveRDS(point_to_point_schedule, "point_to_point_schedule.rds")
cat("Point-to-point schedule saved:", nrow(point_to_point_schedule), "flights\n")
```

Each aircraft for point-to-point carriers is treated as an independent agent traversing a Markov chain of routes. First, it extracts fleet sizes from the observed data by counting unique transponder codes per airline, computes median block times for every route, and builds a transition matrix capturing the probability of flying to each destination given a current airport. Each simulated aircraft starts at a randomly chosen airport weighted by observed departure frequency, then repeatedly samples its next destination from the transition matrix, flies the median block time, turns around in 45 minutes, and continues until the seven-day simulation window expires.

```{r combine-schedules}
schedule <- bind_rows(final_formatted, point_to_point_schedule)
cat("Combined schedule:", nrow(schedule), "flights\n")
```

## Network Simulation

```{r simulation-config}
# # Master configuration object - all parameters in one place
# sim_config <- list(
#   # Time parameters
#   sim_days = 7,
#   time_step_minutes = 1,
#   
# 
#   # Airport capacity (arrivals + departures per hour)
#   capacity = list(
#     large_hub = 90,      # ATL, ORD, DFW, DEN, LAX, etc.
#     medium_hub = 60,     # CLT, PHX, IAH, etc.
#     small_hub = 40,      # BNA, AUS, RDU, etc.
#     non_hub = 20         # Smaller airports
#   ),
#   
# 
#   # Turnaround times (minutes) - fitted from observed data
#   turnaround = list(
#     hub_carrier_at_hub = list(shape = 4.2, rate = 0.08),      # ~52 min mean
#     hub_carrier_at_spoke = list(shape = 3.8, rate = 0.10),    # ~38 min mean
#     p2p_carrier = list(shape = 3.5, rate = 0.09)              # ~39 min mean
#   ),
#   
#   # Minimum connection times (minutes)
#   mct = list(
#     domestic_domestic = 45,
#     domestic_international = 90
#   ),
#   
#   # Delay distributions by source
#   delays = list(
#     # Carrier delays (maintenance, crew, etc.)
#     carrier = list(
#       prob = 0.08,
#       distribution = "lognormal",
#       meanlog = 2.5,
#       sdlog = 0.8
#     ),
#     # Late arriving aircraft (propagated - handled separately)
#     # NAS delays (air traffic control, airport operations)
#     nas = list(
#       prob = 0.05,
#       distribution = "gamma",
#       shape = 2.0,
#       rate = 0.1
#     ),
#     # Security delays
#     security = list(
#       prob = 0.005,
#       distribution = "exponential",
#       rate = 0.05
#     ),
#     # Weather delays (stochastic, localized)
#     weather = list(
#       prob_base = 0.03,
#       distribution = "gamma",
#       shape = 1.5,
#       rate = 0.05
#     )
#   ),
#   
#   # Block time variability (as fraction of scheduled)
#   block_variability = list(
#     mean_factor = 1.0,
#     sd_factor = 0.08
#   ),
#   
#   # Passenger modeling
#   passengers = list(
#     load_factor = 0.85,
#     avg_seats = 150,
#     connecting_pct = 0.30
#   ),
#   
#   # Monte Carlo settings
#   monte_carlo = list(
#     n_replications = 30,
#     seed_base = 12345,
#     parallel_cores = 4
#   )
# )
```

### Airport Classification

```{r airport-classification}
# # Classify airports by traffic volume for capacity assignment
# classify_airports <- function(schedule) {
#   airport_traffic <- schedule |>
#     pivot_longer(cols = c(departure, arrival), names_to = "type", values_to = "airport") |>
#     count(airport, name = "daily_ops") |>
#     mutate(daily_ops = daily_ops / 7) |>  # Average daily
#     arrange(desc(daily_ops))
#   
#   # FAA hub classification thresholds (adjusted for simulation)
#   airport_traffic |>
#     mutate(
#       hub_class = case_when(
#         daily_ops >= 400 ~ "large_hub",
#         daily_ops >= 200 ~ "medium_hub",
#         daily_ops >= 50 ~ "small_hub",
#         TRUE ~ "non_hub"
#       ),
#       capacity_per_hour = case_when(
#         hub_class == "large_hub" ~ sim_config$capacity$large_hub,
#         hub_class == "medium_hub" ~ sim_config$capacity$medium_hub,
#         hub_class == "small_hub" ~ sim_config$capacity$small_hub,
#         TRUE ~ sim_config$capacity$non_hub
#       )
#     )
# }
```

### Delay Generation Functions

```{r delay-generators}
# # Generate delays from fitted distributions
# generate_carrier_delay <- function(n, config = sim_config) {
#   params <- config$delays$carrier
#   delays <- rep(0, n)
#   affected <- runif(n) < params$prob
#   
#   if (sum(affected) > 0) {
#     delays[affected] <- rlnorm(
#       sum(affected),
#       meanlog = params$meanlog,
#       sdlog = params$sdlog
#     )
#   }
#   round(pmin(delays, 300))  # Cap at 5 hours
# }
# 
# generate_nas_delay <- function(n, config = sim_config) {
#   params <- config$delays$nas
#   delays <- rep(0, n)
#   affected <- runif(n) < params$prob
#   
#   if (sum(affected) > 0) {
#     delays[affected] <- rgamma(
#       sum(affected),
#       shape = params$shape,
#       rate = params$rate
#     )
#   }
#   round(pmin(delays, 180))  # Cap at 3 hours
# }
# 
# generate_security_delay <- function(n, config = sim_config) {
#   params <- config$delays$security
#   delays <- rep(0, n)
#   affected <- runif(n) < params$prob
#   
#   if (sum(affected) > 0) {
#     delays[affected] <- rexp(sum(affected), rate = params$rate)
#   }
#   round(pmin(delays, 120))  # Cap at 2 hours
# }
# 
# generate_weather_delay <- function(n, airport_weather_factor, config = sim_config)
# {
#   # airport_weather_factor: multiplier based on current weather conditions
#   params <- config$delays$weather
#   prob_adjusted <- params$prob_base * airport_weather_factor
#   
#   delays <- rep(0, n)
#   affected <- runif(n) < prob_adjusted
#   
#   if (sum(affected) > 0) {
#     delays[affected] <- rgamma(
#       sum(affected),
#       shape = params$shape,
#       rate = params$rate
#     )
#   }
#   round(pmin(delays, 360))  # Cap at 6 hours
# }
# 
# # Generate turnaround time
# generate_turnaround <- function(n, carrier_type, at_hub, config = sim_config) {
#   if (carrier_type == "hub_spoke") {
#     if (at_hub) {
#       params <- config$turnaround$hub_carrier_at_hub
#     } else {
#       params <- config$turnaround$hub_carrier_at_spoke
#     }
#   } else {
#     params <- config$turnaround$p2p_carrier
#   }
#   
#   round(rgamma(n, shape = params$shape, rate = params$rate))
# }
# 
# # Generate block time with variability
# generate_actual_block <- function(scheduled_block, config = sim_config) {
#   n <- length(scheduled_block)
#   factors <- rnorm(
#     n,
#     mean = config$block_variability$mean_factor,
#     sd = config$block_variability$sd_factor
#   )
#   factors <- pmax(0.8, pmin(1.3, factors))  # Bound between 80% and 130%
#   round(scheduled_block * factors)
# }
```

### Weather Scenario Generator

```{r weather-scenarios}
# # Define weather event types
# create_weather_scenario <- function(scenario_type = "clear", 
#                                     affected_airports = NULL,
#                                     start_hour = 0, 
#                                     duration_hours = 24,
#                                     intensity = 1.0) {
#   
#   scenarios <- list(
#     clear = list(
#       name = "Clear conditions",
#       capacity_factor = 1.0,
#       delay_factor = 1.0,
#       ground_stop = FALSE
#     ),
#     thunderstorms = list(
#       name = "Convective weather",
#       capacity_factor = 0.5,
#       delay_factor = 4.0,
#       ground_stop = FALSE
#     ),
#     winter_storm = list(
#       name = "Winter storm",
#       capacity_factor = 0.3,
#       delay_factor = 6.0,
#       ground_stop = FALSE
#     ),
#     ground_stop = list(
#       name = "Ground stop (GDP/GS)",
#       capacity_factor = 0.0,
#       delay_factor = 10.0,
#       ground_stop = TRUE
#     ),
#     low_visibility = list(
#       name = "Low visibility/fog",
#       capacity_factor = 0.6,
#       delay_factor = 2.5,
#       ground_stop = FALSE
#     )
#   )
#   
#   base_scenario <- scenarios[[scenario_type]]
#   
#   # Apply intensity modifier
#   base_scenario$capacity_factor <- base_scenario$capacity_factor^intensity
#   base_scenario$delay_factor <- base_scenario$delay_factor * intensity
#   
#   list(
#     type = scenario_type,
#     params = base_scenario,
#     affected_airports = affected_airports,
#     start_minute = start_hour * 60,
#     end_minute = (start_hour + duration_hours) * 60
#   )
# }
# 
# # Check if airport is affected by weather at given time
# get_weather_impact <- function(airport, minute_of_day, day, weather_events) {
#   impact <- list(capacity_factor = 1.0, delay_factor = 1.0, ground_stop = FALSE)
#   
#   for (event in weather_events) {
#     if (!is.null(event$affected_airports) && 
#         airport %in% event$affected_airports &&
#         minute_of_day >= event$start_minute && 
#         minute_of_day < event$end_minute) {
#       
#       impact$capacity_factor <- min(impact$capacity_factor, event$params$capacity_factor)
#       impact$delay_factor <- max(impact$delay_factor, event$params$delay_factor)
#       impact$ground_stop <- impact$ground_stop || event$params$ground_stop
#     }
#   }
#   
#   impact
# }
```

### Airport Queue Manager

```{r airport-queues}
# # Initialize airport state tracking
# initialize_airport_state <- function(airport_info) {
#   airport_info |>
#     mutate(
#       departure_queue = vector("list", n()),
#       arrival_queue = vector("list", n()),
#       last_departure = 0,
#       last_arrival = 0,
#       departures_this_hour = 0,
#       arrivals_this_hour = 0,
#       current_hour = 0
#     ) |>
#     select(airport, hub_class, capacity_per_hour, 
#            departure_queue, arrival_queue,
#            last_departure, last_arrival,
#            departures_this_hour, arrivals_this_hour,
#            current_hour)
# }
# 
# # Request departure slot from airport
# request_departure_slot <- function(airport_state, airport, requested_time, 
#                                    weather_events, day) {
#   idx <- which(airport_state$airport == airport)
#   if (length(idx) == 0) return(list(granted_time = requested_time, queue_delay = 0))
#   
#   # Get weather impact
#   weather <- get_weather_impact(airport, requested_time %% 1440, day, weather_events)
#   
#   # Ground stop - delay until end
#   if (weather$ground_stop) {
#     for (event in weather_events) {
#       if (!is.null(event$affected_airports) && 
#           airport %in% event$affected_airports &&
#           event$params$ground_stop) {
#         return(list(
#           granted_time = event$end_minute + (day - 1) * 1440,
#           queue_delay = event$end_minute + (day - 1) * 1440 - requested_time
#         ))
#       }
#     }
#   }
#   
#   # Adjust capacity for weather
#   effective_capacity <- airport_state$capacity_per_hour[idx] * weather$capacity_factor
#   min_separation <- 60 / max(1, effective_capacity)  # Minutes between operations
#   
#   # Calculate earliest available slot
#   last_dep <- airport_state$last_departure[idx]
#   earliest_available <- max(requested_time, last_dep + min_separation)
#   
#   queue_delay <- earliest_available - requested_time
#   
#   # Update state
#   airport_state$last_departure[idx] <<- earliest_available
#   
#   list(granted_time = earliest_available, queue_delay = queue_delay)
# }
# 
# # Request arrival slot from airport
# request_arrival_slot <- function(airport_state, airport, requested_time,
#                                  weather_events, day) {
#   idx <- which(airport_state$airport == airport)
#   if (length(idx) == 0) return(list(granted_time = requested_time, queue_delay = 0))
#   
#   weather <- get_weather_impact(airport, requested_time %% 1440, day, weather_events)
#   
#   # Holding patterns for ground stop arrivals
#   if (weather$ground_stop) {
#     for (event in weather_events) {
#       if (!is.null(event$affected_airports) && 
#           airport %in% event$affected_airports &&
#           event$params$ground_stop) {
#         # Add holding delay (randomized)
#         hold_time <- runif(1, 20, 60)
#         return(list(
#           granted_time = requested_time + hold_time,
#           queue_delay = hold_time
#         ))
#       }
#     }
#   }
#   
#   effective_capacity <- airport_state$capacity_per_hour[idx] * weather$capacity_factor
#   min_separation <- 60 / max(1, effective_capacity)
#   
#   last_arr <- airport_state$last_arrival[idx]
#   earliest_available <- max(requested_time, last_arr + min_separation)
#   
#   queue_delay <- earliest_available - requested_time
#   airport_state$last_arrival[idx] <<- earliest_available
#   
#   list(granted_time = earliest_available, queue_delay = queue_delay)
# }
```

### Flight State Machine

```{r flight-state-machine}
# # Define flight states
# FLIGHT_STATES <- c(
#   "SCHEDULED",
#   "BOARDING",
#   "TAXI_OUT",
#   "AIRBORNE",
#   "TAXI_IN",
#   "ARRIVED",
#   "CANCELLED"
# )
# 
# # Initialize flight for simulation
# initialize_flight <- function(flight_row, day_offset = 0) {
#   list(
#     flight_id = paste0(flight_row$callsign, "_D", day_offset, "_", 
#                        format(flight_row$departure_time, "%H%M")),
#     airline = flight_row$airline,
#     callsign = flight_row$callsign,
#     tail_number = flight_row$tail_number,
#     departure = flight_row$departure,
#     arrival = flight_row$arrival,
#     
#     # Scheduled times (in minutes from midnight of day 1)
#     sched_dep_min = as.numeric(flight_row$departure_time) / 60 + day_offset * 1440,
#     sched_arr_min = as.numeric(flight_row$arrival_time) / 60 + day_offset * 1440,
#     sched_block = flight_row$duration / 60,
#     
#     # Actual times (to be computed)
#     actual_dep_min = NA_real_,
#     actual_arr_min = NA_real_,
#     actual_block = NA_real_,
#     
#     # Delay components
#     delay_carrier = 0,
#     delay_weather = 0,
#     delay_nas = 0,
#     delay_security = 0,
#     delay_late_aircraft = 0,
#     delay_queue_dep = 0,
#     delay_queue_arr = 0,
#     
#     # State
#     state = "SCHEDULED",
#     
#     # Passenger info
#     passengers_booked = 0,
#     passengers_connecting = 0,
#     passengers_misconnected = 0
#   )
# }
```

### Core Simulation Engine

```{r simulation-engine}
# # Main simulation function for a single replication
# run_single_simulation <- function(schedule, 
#                                   airport_info,
#                                   weather_events = list(),
#                                   config = sim_config,
#                                   seed = NULL) {
#   
#   if (!is.null(seed)) set.seed(seed)
#   
#   # Classify airlines
#   hub_spoke_carriers <- c("UAL", "DAL", "AAL", "ASA")
#   
#   # Initialize airport state
#   airport_state <- initialize_airport_state(airport_info)
#   
#   # Pre-generate all exogenous delays
#   n_flights <- nrow(schedule)
#   exo_delays <- tibble(
#     carrier = generate_carrier_delay(n_flights, config),
#     nas = generate_nas_delay(n_flights, config),
#     security = generate_security_delay(n_flights, config)
#   )
#   
#   # Initialize all flights
#   days_map <- c("Monday" = 0, "Tuesday" = 1, "Wednesday" = 2, "Thursday" = 3,
#                 "Friday" = 4, "Saturday" = 5, "Sunday" = 6)
#   
#   flights <- schedule |>
#     mutate(day_offset = days_map[as.character(day_of_week)]) |>
#     rowwise() |>
#     mutate(flight_obj = list(initialize_flight(cur_data(), day_offset))) |>
#     ungroup() |>
#     pull(flight_obj)
#   
#   # Assign exogenous delays
#   for (i in seq_along(flights)) {
#     flights[[i]]$delay_carrier <- exo_delays$carrier[i]
#     flights[[i]]$delay_nas <- exo_delays$nas[i]
#     flights[[i]]$delay_security <- exo_delays$security[i]
#     
#     # Generate weather delay based on departure airport conditions
#     weather_impact <- get_weather_impact(
#       flights[[i]]$departure,
#       flights[[i]]$sched_dep_min %% 1440,
#       floor(flights[[i]]$sched_dep_min / 1440) + 1,
#       weather_events
#     )
#     flights[[i]]$delay_weather <- generate_weather_delay(
#       1, weather_impact$delay_factor, config
#     )
#     
#     # Generate actual block time
#     flights[[i]]$actual_block <- generate_actual_block(
#       flights[[i]]$sched_block, config
#     )
#     
#     # Assign passengers
#     avg_pax <- config$passengers$avg_seats * config$passengers$load_factor
#     flights[[i]]$passengers_booked <- round(rnorm(1, avg_pax, avg_pax * 0.1))
#     flights[[i]]$passengers_connecting <- round(
#       flights[[i]]$passengers_booked * config$passengers$connecting_pct
#     )
#   }
#   
#   # Sort by scheduled departure
#   flight_order <- order(sapply(flights, function(f) f$sched_dep_min))
#   flights <- flights[flight_order]
#   
#   # Track aircraft availability: tail_number -> ready_time
#   aircraft_ready <- list()
#   
#   # Process flights in order
#   for (i in seq_along(flights)) {
#     f <- flights[[i]]
#     
#     # Check aircraft propagation delay
#     if (!is.null(aircraft_ready[[f$tail_number]])) {
#       ready_time <- aircraft_ready[[f$tail_number]]$ready_time
#       ready_airport <- aircraft_ready[[f$tail_number]]$airport
#       
#       if (ready_airport == f$departure && ready_time > f$sched_dep_min) {
#         f$delay_late_aircraft <- ready_time - f$sched_dep_min
#       }
#     }
#     
#     # Total departure delay
#     total_dep_delay <- f$delay_carrier + f$delay_weather + f$delay_nas + 
#                         f$delay_security + f$delay_late_aircraft
#     
#     desired_dep <- f$sched_dep_min + total_dep_delay
#     
#     # Request departure slot
#     day <- floor(f$sched_dep_min / 1440) + 1
#     dep_slot <- request_departure_slot(
#       airport_state, f$departure, desired_dep, weather_events, day
#     )
#     
#     f$delay_queue_dep <- dep_slot$queue_delay
#     f$actual_dep_min <- dep_slot$granted_time
#     
#     # Calculate arrival
#     desired_arr <- f$actual_dep_min + f$actual_block
#     
#     # Request arrival slot
#     arr_day <- floor(desired_arr / 1440) + 1
#     arr_slot <- request_arrival_slot(
#       airport_state, f$arrival, desired_arr, weather_events, arr_day
#     )
#     
#     f$delay_queue_arr <- arr_slot$queue_delay
#     f$actual_arr_min <- arr_slot$granted_time
#     
#     # Update state
#     f$state <- "ARRIVED"
#     
#     # Update aircraft availability
#     is_hub_spoke <- f$airline %in% hub_spoke_carriers
#     is_at_hub <- f$arrival %in% c("KATL", "KORD", "KDFW", "KDEN", "KLAX", 
#                                    "KJFK", "KSFO", "KCLT", "KLAS", "KPHX",
#                                    "KMIA", "KIAH", "KEWR", "KMSP", "KBOS",
#                                    "KDTW", "KPHL", "KLGA", "KFLL", "KDCA",
#                                    "KSLC", "KSAN", "KBWI", "KTPA", "KAUS")
#     
#     turn_time <- generate_turnaround(
#       1, 
#       if_else(is_hub_spoke, "hub_spoke", "p2p"),
#       is_at_hub,
#       config
#     )
#     
#     aircraft_ready[[f$tail_number]] <- list(
#       ready_time = f$actual_arr_min + turn_time,
#       airport = f$arrival
#     )
#     
#     flights[[i]] <- f
#   }
#   
#   # Convert results to data frame
#   results <- tibble(
#     flight_id = sapply(flights, function(f) f$flight_id),
#     airline = sapply(flights, function(f) f$airline),
#     tail_number = sapply(flights, function(f) f$tail_number),
#     departure = sapply(flights, function(f) f$departure),
#     arrival = sapply(flights, function(f) f$arrival),
#     
#     sched_dep_min = sapply(flights, function(f) f$sched_dep_min),
#     sched_arr_min = sapply(flights, function(f) f$sched_arr_min),
#     actual_dep_min = sapply(flights, function(f) f$actual_dep_min),
#     actual_arr_min = sapply(flights, function(f) f$actual_arr_min),
#     
#     delay_carrier = sapply(flights, function(f) f$delay_carrier),
#     delay_weather = sapply(flights, function(f) f$delay_weather),
#     delay_nas = sapply(flights, function(f) f$delay_nas),
#     delay_security = sapply(flights, function(f) f$delay_security),
#     delay_late_aircraft = sapply(flights, function(f) f$delay_late_aircraft),
#     delay_queue_dep = sapply(flights, function(f) f$delay_queue_dep),
#     delay_queue_arr = sapply(flights, function(f) f$delay_queue_arr),
#     
#     passengers_booked = sapply(flights, function(f) f$passengers_booked),
#     passengers_connecting = sapply(flights, function(f) f$passengers_connecting)
#   ) |>
#     mutate(
#       dep_delay = actual_dep_min - sched_dep_min,
#       arr_delay = actual_arr_min - sched_arr_min,
#       total_exo_delay = delay_carrier + delay_weather + delay_nas + delay_security,
#       total_prop_delay = delay_late_aircraft + delay_queue_dep + delay_queue_arr
#     )
#   
#   results
# }
```

### Passenger Connection Analysis

```{r passenger-connections}
# # Analyze passenger misconnections
# analyze_connections <- function(sim_results, mct = 45) {
#   # Build potential connections: arrive at airport, depart from same airport
#   arrivals <- sim_results |>
#     select(inbound_flight = flight_id, connect_airport = arrival,
#            inbound_arr = actual_arr_min, pax_connecting = passengers_connecting)
#   
#   departures <- sim_results |>
#     select(outbound_flight = flight_id, connect_airport = departure,
#            outbound_dep = actual_dep_min)
#   
#   # Find all potential connections (same airport, reasonable time window)
#   connections <- arrivals |>
#     inner_join(departures, by = "connect_airport", relationship = "many-to-many") |>
#     filter(
#       inbound_flight != outbound_flight,
#       outbound_dep > inbound_arr,
#       outbound_dep - inbound_arr <= 240  # Max 4 hour connection
#     ) |>
#     mutate(
#       connection_time = outbound_dep - inbound_arr,
#       made_connection = connection_time >= mct
#     )
#   
#   # Calculate misconnection rate
#   misconnections <- connections |>
#     filter(!made_connection) |>
#     summarise(
#       total_misconnections = n(),
#       pax_affected = sum(pax_connecting * 0.1, na.rm = TRUE)  # Assume 10% try this connection
#     )
#   
#   list(
#     connections = connections,
#     misconnections = misconnections,
#     misconnection_rate = mean(!connections$made_connection)
#   )
# }
# ```
# 
# ### Monte Carlo Framework
# 
# ```{r monte-carlo}
# # Run Monte Carlo simulation
# run_monte_carlo <- function(schedule, 
#                             airport_info,
#                             weather_events = list(),
#                             n_replications = 30,
#                             config = sim_config,
#                             parallel = TRUE) {
#   
#   if (parallel && config$monte_carlo$parallel_cores > 1) {
#     cl <- makeCluster(min(config$monte_carlo$parallel_cores, detectCores() - 1))
#     registerDoParallel(cl)
#     
#     results <- foreach(
#       rep = 1:n_replications,
#       .combine = bind_rows,
#       .packages = c("tidyverse", "truncnorm"),
#       .export = c("run_single_simulation", "initialize_airport_state",
#                   "initialize_flight", "generate_carrier_delay",
#                   "generate_nas_delay", "generate_security_delay",
#                   "generate_weather_delay", "generate_turnaround",
#                   "generate_actual_block", "request_departure_slot",
#                   "request_arrival_slot", "get_weather_impact",
#                   "sim_config")
#     ) %dopar% {
#       seed <- config$monte_carlo$seed_base + rep
#       sim_result <- run_single_simulation(
#         schedule, airport_info, weather_events, config, seed
#       )
#       sim_result$replication <- rep
#       sim_result
#     }
#     
#     stopCluster(cl)
#   } else {
#     results <- map_dfr(1:n_replications, function(rep) {
#       seed <- config$monte_carlo$seed_base + rep
#       sim_result <- run_single_simulation(
#         schedule, airport_info, weather_events, config, seed
#       )
#       sim_result$replication <- rep
#       sim_result
#     })
#   }
#   
#   results
# }
# 
# # Summarize Monte Carlo results
# summarize_monte_carlo <- function(mc_results) {
#   # Overall metrics
#   overall <- mc_results |>
#     group_by(replication) |>
#     summarise(
#       n_flights = n(),
#       mean_dep_delay = mean(dep_delay),
#       mean_arr_delay = mean(arr_delay),
#       median_arr_delay = median(arr_delay),
#       pct_delayed_15 = mean(arr_delay > 15) * 100,
#       pct_delayed_60 = mean(arr_delay > 60) * 100,
#       total_delay_minutes = sum(arr_delay),
#       prop_delay_share = sum(total_prop_delay) / sum(pmax(arr_delay, 0.1)),
#       .groups = "drop"
#     )
#   
#   # Confidence intervals
#   ci <- overall |>
#     summarise(
#       across(
#         c(mean_arr_delay, pct_delayed_15, prop_delay_share),
#         list(
#           mean = ~mean(.),
#           lower = ~quantile(., 0.025),
#           upper = ~quantile(., 0.975)
#         )
#       )
#     )
#   
#   # By airline
#   by_airline <- mc_results |>
#     group_by(airline, replication) |>
#     summarise(
#       mean_arr_delay = mean(arr_delay),
#       pct_delayed_15 = mean(arr_delay > 15) * 100,
#       .groups = "drop"
#     ) |>
#     group_by(airline) |>
#     summarise(
#       delay_mean = mean(mean_arr_delay),
#       delay_lower = quantile(mean_arr_delay, 0.025),
#       delay_upper = quantile(mean_arr_delay, 0.975),
#       .groups = "drop"
#     )
#   
#   # By airport
#   by_airport <- mc_results |>
#     pivot_longer(cols = c(departure, arrival), names_to = "type", 
#                  values_to = "airport") |>
#     group_by(airport, replication) |>
#     summarise(mean_delay = mean(arr_delay), .groups = "drop") |>
#     group_by(airport) |>
#     summarise(
#       delay_mean = mean(mean_delay),
#       delay_sd = sd(mean_delay),
#       .groups = "drop"
#     ) |>
#     arrange(desc(delay_mean))
#   
#   list(
#     overall = overall,
#     confidence_intervals = ci,
#     by_airline = by_airline,
#     by_airport = by_airport
#   )
# }
```

### Validation Framework

```{r validation}
# # Compare simulation to observed data
# validate_simulation <- function(sim_results, observed_flights) {
#   # Match flights between simulated and observed
#   # Using departure airport, arrival airport, and approximate time
#   
#   sim_summary <- sim_results |>
#     group_by(departure, arrival) |>
#     summarise(
#       sim_n = n(),
#       sim_mean_delay = mean(arr_delay),
#       sim_sd_delay = sd(arr_delay),
#       sim_pct_delayed = mean(arr_delay > 15) * 100,
#       .groups = "drop"
#     )
#   
#   # Calculate observed metrics from original data
#   obs_summary <- observed_flights |>
#     mutate(
#       duration_actual = as.numeric(difftime(lastseen, firstseen, units = "mins"))
#     ) |>
#     group_by(departure, arrival) |>
#     summarise(
#       obs_n = n(),
#       obs_mean_duration = mean(duration_actual, na.rm = TRUE),
#       .groups = "drop"
#     )
#   
#   # Join and compare
#   comparison <- sim_summary |>
#     inner_join(obs_summary, by = c("departure", "arrival")) |>
#     mutate(
#       volume_ratio = sim_n / obs_n
#     )
#   
#   # Statistical tests
#   validation_metrics <- list(
#     route_coverage = nrow(comparison) / nrow(sim_summary),
#     volume_correlation = cor(comparison$sim_n, comparison$obs_n, 
#                              use = "complete.obs"),
#     mean_volume_ratio = mean(comparison$volume_ratio, na.rm = TRUE)
#   )
#   
#   list(
#     comparison = comparison,
#     metrics = validation_metrics
#   )
# }
```

### Sensitivity Analysis

```{r sensitivity}
# # Run sensitivity analysis on key parameters
# run_sensitivity_analysis <- function(schedule, airport_info, 
#                                      base_config = sim_config,
#                                      n_reps = 10) {
#   
#   # Parameters to vary
#   sensitivity_params <- list(
#     carrier_delay_prob = c(0.04, 0.08, 0.12, 0.16),
#     nas_delay_prob = c(0.025, 0.05, 0.075, 0.10),
#     turnaround_rate = c(0.06, 0.08, 0.10, 0.12),
#     capacity_factor = c(0.8, 1.0, 1.2, 1.4)
#   )
#   
#   results <- list()
#   
#   # Vary carrier delay probability
#   cat("Testing carrier delay probability sensitivity...\n")
#   carrier_results <- map_dfr(sensitivity_params$carrier_delay_prob, function(p) {
#     config_mod <- base_config
#     config_mod$delays$carrier$prob <- p
#     
#     sim <- run_monte_carlo(schedule, airport_info, list(), n_reps, 
#                            config_mod, parallel = FALSE)
#     
#     sim |>
#       summarise(
#         param_value = p,
#         mean_arr_delay = mean(arr_delay),
#         pct_delayed_15 = mean(arr_delay > 15) * 100
#       )
#   })
#   results$carrier_delay <- carrier_results
#   
#   # Vary NAS delay probability
#   cat("Testing NAS delay probability sensitivity...\n")
#   nas_results <- map_dfr(sensitivity_params$nas_delay_prob, function(p) {
#     config_mod <- base_config
#     config_mod$delays$nas$prob <- p
#     
#     sim <- run_monte_carlo(schedule, airport_info, list(), n_reps,
#                            config_mod, parallel = FALSE)
#     
#     sim |>
#       summarise(
#         param_value = p,
#         mean_arr_delay = mean(arr_delay),
#         pct_delayed_15 = mean(arr_delay > 15) * 100
#       )
#   })
#   results$nas_delay <- nas_results
#   
#   # Vary capacity
#   cat("Testing capacity sensitivity...\n")
#   cap_results <- map_dfr(sensitivity_params$capacity_factor, function(f) {
#     airport_info_mod <- airport_info |>
#       mutate(capacity_per_hour = round(capacity_per_hour * f))
#     
#     sim <- run_monte_carlo(schedule, airport_info_mod, list(), n_reps,
#                            base_config, parallel = FALSE)
#     
#     sim |>
#       summarise(
#         param_value = f,
#         mean_arr_delay = mean(arr_delay),
#         pct_delayed_15 = mean(arr_delay > 15) * 100
#       )
#   })
#   results$capacity <- cap_results
#   
#   results
# }
```

### Visualization Functions

```{r visualization}
# # Plot delay distribution by source
# plot_delay_sources <- function(sim_results) {
#   delay_sources <- sim_results |>
#     summarise(
#       Carrier = sum(delay_carrier),
#       Weather = sum(delay_weather),
#       NAS = sum(delay_nas),
#       Security = sum(delay_security),
#       `Late Aircraft` = sum(delay_late_aircraft),
#       `Queue (Dep)` = sum(delay_queue_dep),
#       `Queue (Arr)` = sum(delay_queue_arr)
#     ) |>
#     pivot_longer(everything(), names_to = "Source", values_to = "Minutes")
#   
#   ggplot(delay_sources, aes(x = reorder(Source, -Minutes), y = Minutes / 1000)) +
#     geom_col(fill = "steelblue") +
#     labs(
#       title = "Total Delay by Source",
#       x = "Delay Source",
#       y = "Total Delay (thousands of minutes)"
#     ) +
#     theme_minimal() +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))
# }
# 
# # Plot propagation cascade
# plot_propagation_cascade <- function(sim_results, sample_tail) {
#   cascade_data <- sim_results |>
#     filter(tail_number == sample_tail) |>
#     arrange(sched_dep_min) |>
#     mutate(
#       leg = row_number(),
#       route = paste(departure, "→", arrival)
#     )
#   
#   ggplot(cascade_data) +
#     geom_segment(
#       aes(x = sched_dep_min / 60, xend = sched_arr_min / 60,
#           y = leg, yend = leg),
#       color = "gray70", linewidth = 3, alpha = 0.5
#     ) +
#     geom_segment(
#       aes(x = actual_dep_min / 60, xend = actual_arr_min / 60,
#           y = leg, yend = leg, color = arr_delay),
#       linewidth = 2
#     ) +
#     geom_point(aes(x = actual_dep_min / 60, y = leg), size = 2) +
#     geom_point(aes(x = actual_arr_min / 60, y = leg), size = 2) +
#     scale_color_gradient2(
#       low = "darkgreen", mid = "yellow", high = "red",
#       midpoint = 15, name = "Arrival\nDelay (min)"
#     ) +
#     scale_y_reverse(breaks = cascade_data$leg, labels = cascade_data$route) +
#     labs(
#       title = paste("Delay Propagation for Aircraft:", sample_tail),
#       x = "Hour of Simulation",
#       y = "Flight Leg"
#     ) +
#     theme_minimal()
# }
# 
# # Plot Monte Carlo convergence
# plot_mc_convergence <- function(mc_summary) {
#   ggplot(mc_summary$overall, aes(x = replication, y = mean_arr_delay)) +
#     geom_line(color = "steelblue") +
#     geom_point(size = 2) +
#     geom_hline(
#       yintercept = mean(mc_summary$overall$mean_arr_delay),
#       linetype = "dashed", color = "red"
#     ) +
#     geom_ribbon(
#       aes(ymin = mc_summary$confidence_intervals$mean_arr_delay_lower,
#           ymax = mc_summary$confidence_intervals$mean_arr_delay_upper),
#       alpha = 0.2, fill = "red"
#     ) +
#     labs(
#       title = "Monte Carlo Convergence: Mean Arrival Delay",
#       x = "Replication",
#       y = "Mean Arrival Delay (minutes)"
#     ) +
#     theme_minimal()
# }
# 
# # Plot delay distribution comparison
# plot_delay_distribution <- function(sim_results) {
#   sim_results |>
#     mutate(
#       carrier_type = if_else(airline %in% c("UAL", "DAL", "AAL", "ASA"),
#                              "Hub-and-Spoke", "Point-to-Point")
#     ) |>
#     ggplot(aes(x = arr_delay, fill = carrier_type)) +
#     geom_density(alpha = 0.5) +
#     geom_vline(xintercept = 0, linetype = "dashed") +
#     geom_vline(xintercept = 15, linetype = "dotted", color = "red") +
#     scale_x_continuous(limits = c(-30, 120)) +
#     labs(
#       title = "Arrival Delay Distribution by Carrier Type",
#       x = "Arrival Delay (minutes)",
#       y = "Density",
#       fill = "Carrier Type"
#     ) +
#     theme_minimal()
# }
# 
# # Plot airport delay heatmap
# plot_airport_heatmap <- function(sim_results, top_n = 20) {
#   airport_delays <- sim_results |>
#     group_by(departure) |>
#     summarise(
#       mean_dep_delay = mean(dep_delay),
#       n_flights = n(),
#       .groups = "drop"
#     ) |>
#     arrange(desc(n_flights)) |>
#     slice_head(n = top_n)
#   
#   airport_order <- airport_delays$departure
#   
#   hourly_delays <- sim_results |>
#     filter(departure %in% airport_order) |>
#     mutate(
#       hour = floor((sched_dep_min %% 1440) / 60),
#       departure = factor(departure, levels = airport_order)
#     ) |>
#     group_by(departure, hour) |>
#     summarise(mean_delay = mean(dep_delay), .groups = "drop")
#   
#   ggplot(hourly_delays, aes(x = hour, y = departure, fill = mean_delay)) +
#     geom_tile() +
#     scale_fill_gradient2(
#       low = "darkgreen", mid = "yellow", high = "red",
#       midpoint = 15, name = "Mean\nDelay"
#     ) +
#     scale_x_continuous(breaks = seq(0, 23, 3)) +
#     labs(
#       title = "Departure Delays by Airport and Hour",
#       x = "Hour of Day",
#       y = "Airport"
#     ) +
#     theme_minimal()
# }
# 
# # Sensitivity analysis plots
# plot_sensitivity <- function(sensitivity_results) {
#   bind_rows(
#     sensitivity_results$carrier_delay |> mutate(param = "Carrier Delay Prob"),
#     sensitivity_results$nas_delay |> mutate(param = "NAS Delay Prob"),
#     sensitivity_results$capacity |> mutate(param = "Capacity Factor")
#   ) |>
#     ggplot(aes(x = param_value, y = mean_arr_delay)) +
#     geom_line(linewidth = 1) +
#     geom_point(size = 3) +
#     facet_wrap(~param, scales = "free_x") +
#     labs(
#       title = "Sensitivity Analysis: Effect on Mean Arrival Delay",
#       x = "Parameter Value",
#       y = "Mean Arrival Delay (minutes)"
#     ) +
#     theme_minimal()
# }
```

## Running the Simulation

```{r run-example, eval=FALSE}
# # Load your schedule data
# schedule <- readRDS("hub_spoke_schedule.rds") |>
#   bind_rows(readRDS("point_to_point_schedule.rds"))
# 
# # Classify airports
# airport_info <- classify_airports(schedule)
# 
# # Baseline simulation (no weather disruption)
# cat("Running baseline simulation...\n")
# baseline_results <- run_monte_carlo(
#   schedule, airport_info, 
#   weather_events = list(),
#   n_replications = 30,
#   parallel = TRUE
# )
# 
# baseline_summary <- summarize_monte_carlo(baseline_results)
# print(baseline_summary$confidence_intervals)
# 
# # Weather scenario: Thunderstorms at major hubs
# cat("\nRunning weather disruption scenario...\n")
# weather_scenario <- list(
#   create_weather_scenario(
#     "thunderstorms",
#     affected_airports = c("KATL", "KORD", "KDFW"),
#     start_hour = 14,
#     duration_hours = 4,
#     intensity = 1.2
#   )
# )
# 
# weather_results <- run_monte_carlo(
#   schedule, airport_info,
#   weather_events = weather_scenario,
#   n_replications = 30,
#   parallel = TRUE
# )
# 
# weather_summary <- summarize_monte_carlo(weather_results)
# 
# # Compare scenarios
# comparison <- tibble(
#   Scenario = c("Baseline", "Thunderstorms"),
#   `Mean Delay` = c(
#     baseline_summary$confidence_intervals$mean_arr_delay_mean,
#     weather_summary$confidence_intervals$mean_arr_delay_mean
#   ),
#   `% Delayed 15+` = c(
#     baseline_summary$confidence_intervals$pct_delayed_15_mean,
#     weather_summary$confidence_intervals$pct_delayed_15_mean
#   ),
#   `Propagation Share` = c(
#     baseline_summary$confidence_intervals$prop_delay_share_mean,
#     weather_summary$confidence_intervals$prop_delay_share_mean
#   )
# )
# 
# print(comparison)
# 
# # Visualizations
# plot_delay_sources(baseline_results |> filter(replication == 1))
# plot_delay_distribution(baseline_results |> filter(replication == 1))
# plot_airport_heatmap(baseline_results |> filter(replication == 1))
# 
# # Sensitivity analysis
# cat("\nRunning sensitivity analysis...\n")
# sensitivity <- run_sensitivity_analysis(
#   schedule |> sample_frac(0.2),  # Use subset for speed
#   airport_info,
#   n_reps = 5
# )
# plot_sensitivity(sensitivity)
```

### Key Improvements Over Original Simulation

Airport capacity - Queue-based with hourly limits
Delay sources - 5 distinct sources with fitted distributions
Weather - Full scenario system with capacity impacts
Turnaround - Gamma-distributed by carrier/location type
Passenger connections - Misconnection analysis
Statistical validity - Monte Carlo with confidence intervals
Validation - Framework for comparing to observed data
Sensitivity - Multi-parameter analysis

## References

