---
title: "Combining for Simulation"
format: html
editor: visual
---

```{r load-packages, message = FALSE}

library(tidyverse)
library(airportr)
library(gganimate)
library(ggimage)
library(maps)
library(gifski)
library(geosphere)
library(httr)
library(jsonlite)
library(fuzzyjoin)
library(rvest)
library(lutz)
library(hms)
library(mgcv)
library(slider)
library(mapproj)
```

### Why Care?

The country’s air transportation is highly interconnected, but thus vulnerable to disruptions. These disruptions lead to delays that expand across the network and affect airlines, airports, and passengers in complex ways. Each day, the United States has 2 million enplanements—which means the impact of even small disturbances can be widespread.

This project aims to build a simulation that can model such disruptions at scale, thinking from a point of view of of robustness, recovery, and systemic impacts—with a final goal of understanding what airport as a hub can help mitigate some of these issues.

Air traffic disruptions have been studied using network theory, agent-based modeling, and operational research. These have shown that disruptions can cascade rapidly. Airlines, airports, and passengers all make interdependent decisions during disruptions, which makes modeling challenging. Existing data sources provide partial insights into these, but using more data and creating a clearer way to simulate is a research gap that can be filled and further used to expand the field of research in aviation.

### Data

Data was built in a few ways. The OpenSky data can be pulled to give last locations of current flights. Can also use Trino to pull previous flights. Original creation of the thinking behind the simulation was done with synthetic data. The decision on less accurate data came from the cost of real scheduling data.

### General Data Cleaning

```{r commercial-airports}

# Getting Commercial Airports for Simulation
us_airports <- 
  readr::read_csv("https://davidmegginson.github.io/ourairports-data/airports.csv") |>
  filter(
    iso_country == "US",
    scheduled_service == "yes",
    type %in% c("large_airport", "medium_airport"),
    !is.na(ident)
  ) |>
  transmute(
    airport_name = name,
    ICAO = stringr::str_to_upper(stringr::str_trim(ident)),
    IATA = stringr::str_to_upper(stringr::str_trim(iata_code)),
    latitude_deg, longitude_deg
  )
```

After pulling from OpenSky, I needed to do some cleaning of the data. First was making sure the airports were correct. Since OpenSky uses the transponder location instead of the schedule data. Some replacement of airports needed to be done and I only used airports from OurAirports that were sized as large or medium with scheduled service—after checking the small airports and determining to not be relevant airports for the airlines being used.

```{r}

us_states <- map_data("state")

us_airports |>
  filter(latitude_deg > 24 & latitude_deg < 50, longitude_deg > -125 & longitude_deg < -66) |>
  ggplot() +
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white") +
  geom_point(aes(x = longitude_deg, y = latitude_deg), 
             size = 1.5) +
  coord_map() + 
  theme_void() +
  labs(title = "Scheduled Service Airports (US Lower 48)")

us_airports <- us_airports |>
  select(-latitude_deg, -longitude_deg)
```

```{r load-data, warning = FALSE}

path <- "opensky_daily_exports"

csv_files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)

data_list <- lapply(csv_files, read.csv)

all_data <- do.call(rbind, data_list)
```

```{r remove-twin-aisles}

planes <- 
  read_csv("https://s3.opensky-network.org/data-samples/metadata/aircraft-database-complete-2025-02.csv") |>
  filter(`'icao24'` %in% all_data$icao24)

twin_aisle_check <- all_data |>
  filter(!icao24 %in% planes$`'icao24'`)

twin_aisles <- planes |>
  filter(str_detect(`'model'`, str_c(c("767", "777", "787", "A330", "A350", "310R", "A310"), collapse = "|")))

MAX_RANGE_NM <- 3500

twin_aisle_flights <- twin_aisle_check |>
  left_join(
    airportr::airports |>
      dplyr::select(ICAO, Latitude, Longitude) |>
      rename(lat = Latitude, lon = Longitude) |>
      rename(dep_latitude = lat, dep_longitude = lon), 
    by = c("departure" = "ICAO")
    ) |>
  left_join(
    airportr::airports |>
      dplyr::select(ICAO, Latitude, Longitude) |>
      rename(lat = Latitude, lon = Longitude) |>
      rename(arr_latitude = lat, arr_longitude = lon), 
    by = c("arrival" = "ICAO")
    ) |>
  mutate(
    distance_m = geosphere::distHaversine(
      cbind(dep_longitude, dep_latitude), 
      cbind(arr_longitude, arr_latitude)), 
    distance_nm = distance_m / 1852
    ) |>
  group_by(icao24) |>
  summarize(over_range = any(distance_m > MAX_RANGE_NM * 1852, na.rm = TRUE), .groups = "drop") |>
  filter(over_range) |>
  pull(icao24)
```

I removed twin aisle jets from the data of airplanes, seeing as most of them do travel outside the contiguous United States and the analysis was focused on this area. OpenSky supplies a Complete Aircraft Database that allowed for checking the hexadecimal codes for airplanes to their ownership and type. For other planes not in this data set, a check of if they completed a flight during the period which was over 3500 nautical miles (the current longest single aisle jet flight operated by US airlines—EWR to ARN by United Airlines is just under this threshold).

```{r}

all_twin_ids <- unique(c(twin_aisles$`'icao24'`, twin_aisle_flights))

# 2. Filter data and join coordinates
map_data <- all_data |>
  filter(icao24 %in% all_twin_ids) |>
  # Join Departure Coords
  left_join(
    airportr::airports |> select(ICAO, dep_lat = Latitude, dep_lon = Longitude),
    by = c("departure" = "ICAO")
  ) |>
  # Join Arrival Coords
  left_join(
    airportr::airports |> select(ICAO, arr_lat = Latitude, arr_lon = Longitude),
    by = c("arrival" = "ICAO")
  ) |>
  # REMOVE NAs AND IDENTICAL POINTS (This fixes the error)
  filter(
    !is.na(dep_lat), !is.na(dep_lon), !is.na(arr_lat), !is.na(arr_lon),
    !(dep_lat == arr_lat & dep_lon == arr_lon) 
  )

# 3. Plot the map
world_map <- map_data("world")

ggplot() +
  # Background World Map (Updated 'size' to 'linewidth')
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
               fill = "gray90", color = "white", linewidth = 0.2) +
  # Flight paths
  geom_curve(data = map_data, 
             aes(x = dep_lon, y = dep_lat, xend = arr_lon, yend = arr_lat),
             alpha = 0.2, linewidth = 0.2, curvature = 0.2) +
  coord_fixed(1.3) +
  theme_void() +
  labs(
    title = "Routes Flown by Twin Aisle",
    subtitle = paste("Total flights:", nrow(map_data))
  )

map_data <- all_data |>
  filter(icao24 %in% all_twin_ids) |>
  left_join(
    airportr::airports |> select(ICAO, dep_lat = Latitude, dep_lon = Longitude),
    by = c("departure" = "ICAO")
  ) |>
  left_join(
    airportr::airports |> select(ICAO, arr_lat = Latitude, arr_lon = Longitude),
    by = c("arrival" = "ICAO")
  ) |>
  filter(
    !is.na(dep_lat), !is.na(dep_lon), !is.na(arr_lat), !is.na(arr_lon),
    !(dep_lat == arr_lat & dep_lon == arr_lon),
    # US Mainland Bounding Box Filters
    dep_lon > -126, dep_lon < -66, dep_lat > 24, dep_lat < 50,
    arr_lon > -126, arr_lon < -66, arr_lat > 24, arr_lat < 50
  )

ggplot() +
  # US States Background
  geom_polygon(data = us_states, aes(x = long, y = lat, group = group), 
               fill = "gray95", color = "white", linewidth = 0.2) +
  # Flight paths
  geom_curve(data = map_data, 
             aes(x = dep_lon, y = dep_lat, xend = arr_lon, yend = arr_lat),
             alpha = 0.3, linewidth = 0.4, curvature = 0.2) +
  coord_fixed(1.3) +
  theme_void() +
  labs(
    title = "Domestic Twin Aisle Flights",
    subtitle = paste("Total flights:", nrow(map_data))
  )

all_data <- all_data |>
  arrange(icao24) |>
  filter(!icao24 %in% twin_aisle_flights & !icao24 %in% twin_aisles$`'icao24'`)
```

```{r replacing-non-commercials, message = FALSE}

airport_lookup_1 <- read_csv("https://davidmegginson.github.io/ourairports-data/airports.csv") |>
  dplyr::select("ICAO" = "ident",  "latitude" = "latitude_deg", "longitude" = "longitude_deg")

download.file("https://nfdc.faa.gov/webContent/28DaySub/extra/22_Jan_2026_APT_CSV.zip", destfile = "APT_CSV.zip", mode = "wb")

contents <- unzip("APT_CSV.zip", list = TRUE)

unzip("APT_CSV.zip", files = contents[grepl("APT_BASE\\.csv$", contents$Name, ignore.case = TRUE), , drop = FALSE]$Name, 
      exdir = "apt_extract")

airport_lookup_2 <- read_csv(list.files("apt_extract", pattern = "APT_BASE\\.csv$", recursive = TRUE, full.names = TRUE, 
                                        ignore.case = TRUE)[1]) |>
  dplyr::select("ICAO" = "ARPT_ID", "latitude" = "LAT_DECIMAL", "longitude" = "LONG_DECIMAL")

airport_lookups <- bind_rows(airport_lookup_1, airport_lookup_2) |>
  distinct(ICAO, .keep_all = TRUE)

commercial_coords <- airportr::airports |>
  dplyr::select(ICAO, lat = Latitude, lon = Longitude) |>
  mutate(ICAO = str_to_upper(str_trim(ICAO))) |>
  distinct() |>
  semi_join(tibble(ICAO = str_to_upper(str_trim(us_airports$ICAO))), by = "ICAO") |>
  filter(!is.na(lat), !is.na(lon))

lookup_coords <- airport_lookups |>
  transmute(from_icao = str_to_upper(str_trim(ICAO)), lat = as.numeric(latitude), lon = as.numeric(longitude)) |>
  distinct(from_icao, .keep_all = TRUE) |>
  filter(!is.na(from_icao), !is.na(lat), !is.na(lon))

nearest_commercial <- function(lat, lon, comm_tbl, limit_meters = 40000) {
  d_m <- geosphere::distHaversine(cbind(comm_tbl$lon, comm_tbl$lat), c(lon, lat))
  
  min_dist <- min(d_m)
  i <- which.min(d_m)
  
  if (min_dist <= limit_meters) {
    tibble(nearest_icao = comm_tbl$ICAO[i], dist_m = min_dist)
  } 
  
  else {
    tibble(nearest_icao = NA_character_, dist_m = min_dist)
  }
}

closest_map_messy <- lookup_coords |>
  mutate(nn = purrr::pmap(list(lat, lon), \(lat, lon)
    nearest_commercial(lat, lon, commercial_coords, limit_meters = 40000)
  )) |>
  unnest(nn) |>
  filter(!from_icao %in% commercial_coords$ICAO, !is.na(nearest_icao))
```

The replacement was done with a simple closest airport calculation for the non-OurAirports ICAOs. In addition to using the OurAirports, the Federal Aviation Administration’s Airport Master Record has location about some of the airfields. The longitude and latitudes were calculated to those in OurAirports, and then setting a limit to within 40 kilometers.

```{r}


# Visualization
target_hubs <- c("KBDL", "KHVN", "KORH", "KHYA", "KEWB", "KBOS", "KBED", 
                 "KACK", "KPVD", "KWST", "KLEB", "KMHT", "KPSM", "KBTV", 
                 "KRUT", "KAUG", "KBGR", "KBHB", "KPQI", "KPWM", "KRKD")

plot_data <- closest_map_messy |>
  filter(nearest_icao %in% target_hubs) |>
  rename(from_lat = lat, from_lon = lon) |>
  
  left_join(commercial_coords, by = c("nearest_icao" = "ICAO")) |>
  rename(to_lat = lat, to_lon = lon)

ne_state_names <- c("massachusetts", "connecticut", "rhode island", 
                    "new hampshire", "vermont", "maine")
ne_map_bg <- map_data("state", region = ne_state_names)

ggplot() +
  geom_polygon(data = ne_map_bg, aes(x = long, y = lat, group = group), 
               fill = "gray95", color = "gray60") +
  
  geom_segment(data = plot_data, 
               aes(x = from_lon, y = from_lat, xend = to_lon, yend = to_lat),
               color = "steelblue", alpha = 0.4, linewidth = 0.3) +
  
  # Small Airports (Sources)
  geom_point(data = plot_data, aes(x = from_lon, y = from_lat), 
             color = "steelblue", size = 0.5, alpha = 0.6) +
  
  # Commercial Hubs (Destinations)
  geom_point(data = plot_data |> distinct(nearest_icao, to_lon, to_lat), 
             aes(x = to_lon, y = to_lat), 
             color = "black", size = 2) +
  coord_map() +
  theme_void() +
  labs(title = "Airport Catchment Areas",
       subtitle = "Mappings to New England Airports"
       )
```

```{r cleaning-airlines-airports}

all_data <- all_data |>
  left_join(closest_map_messy, by = c("departure" = "from_icao")) |>
  mutate(departure = if_else(!is.na(nearest_icao), nearest_icao, departure)) |>
  dplyr::select(-nearest_icao) |>
  left_join(closest_map_messy, by = c("arrival" = "from_icao")) |>
  mutate(arrival = if_else(!is.na(nearest_icao), nearest_icao, arrival))

flights <- all_data |>
  filter(
    departure %in% us_airports$ICAO, 
    arrival %in% us_airports$ICAO, 
    departure != arrival,
    str_sub(callsign, 1, 3) %in% c("AAL", "ASA", "DAL", "EDV", "FFT", "JBU", "NKS", "RPA", 
                                   "SKW", "SWA", "UAL", "QXE", "JIA", "ENY", "PDT")) |>
  mutate(
    firstseen = ymd_hms(firstseen),
    lastseen = ymd_hms(lastseen),
    duration = as.numeric(difftime(lastseen, firstseen, units = "secs")) + 30 * 60,
    lastseen = firstseen + duration,
    route_from = pmin(departure, arrival),
    route_to = pmax(departure, arrival)
  )
```

```{r add-timezones, warning = FALSE}

airports_ref <- airportr::airports |>
  dplyr::select(ICAO, lat = Latitude, lon = Longitude) |>
  distinct()

tz_lookup <- airports_ref |>
  rowwise() |>
  mutate(timezone = tz_lookup_coords(lat, lon, method = "fast")) |>
  ungroup()

flights <- flights |>
  mutate(
    firstseen = lubridate::ymd_hms(firstseen, tz = "UTC"),
    lastseen  = lubridate::ymd_hms(lastseen,  tz = "UTC")
  )

flights <- flights |>
  dplyr::select(-any_of(c("departure_tz", "arrival_tz", "departure_tz.x", "departure_tz.y", "arrival_tz.x", "arrival_tz.y", 
                          "dep_local", "arr_local", "dep_minute", "arr_minute", "dep_hour", "arr_hour")))

dep_tz_tbl <- tz_lookup |>
  dplyr::select(ICAO, departure_tz = timezone)

arr_tz_tbl <- tz_lookup |>
  dplyr::select(ICAO, arrival_tz = timezone)

flights <- flights |>
  left_join(dep_tz_tbl, by = c("departure" = "ICAO")) |>
  left_join(arr_tz_tbl, by = c("arrival" = "ICAO")) |>
  mutate(
    firstseen = as.POSIXct(firstseen, tz = "UTC"),
    lastseen  = as.POSIXct(lastseen,  tz = "UTC"),

    departure_tz = departure_tz |> as.character() |> stringr::str_trim(),
    arrival_tz   = arrival_tz   |> as.character() |> stringr::str_trim(),

    departure_tz = dplyr::if_else(is.na(departure_tz) | !departure_tz %in% OlsonNames(), "UTC", departure_tz),
    arrival_tz   = dplyr::if_else(is.na(arrival_tz)   | !arrival_tz   %in% OlsonNames(), "UTC", arrival_tz)
  ) |>
  group_by(departure_tz) |>
  mutate(dep_local = lubridate::with_tz(firstseen, unique(departure_tz))) |>
  ungroup() |>
  group_by(arrival_tz) |>
  mutate(arr_local = lubridate::with_tz(lastseen, unique(arrival_tz))) |>
  ungroup() |>
  mutate(
    dep_minute = lubridate::hour(dep_local) * 60 + lubridate::minute(dep_local),
    arr_minute = lubridate::hour(arr_local) * 60 + lubridate::minute(arr_local),
    dep_hour   = lubridate::hour(dep_local),
    arr_hour   = lubridate::hour(arr_local)
  )
```

```{r turnaround-distribution}

turnaround_times <- flights |>
  arrange(icao24, firstseen) |>
  group_by(icao24) |>
  mutate(turnaround_mins = (as.numeric(firstseen) - lag(as.numeric(lastseen))) / 60) |>
  ungroup() |>
  filter(!is.na(turnaround_mins), turnaround_mins > 0, turnaround_mins < 120) |>
  pull(turnaround_mins)

fit <- MASS::fitdistr(turnaround_times, densfun = "gamma")
shape_hat <- unname(fit$estimate["shape"])
rate_hat  <- unname(fit$estimate["rate"])

shape_hat
rate_hat
```

### Hub-and-Spoke Carriers

For hub and spoke airlines, some key ideas are the hubs they have, the banks they operate at airports, and destinations flown to.

```{r select-hubs}

target_airlines <- c("UAL", "DAL", "AAL", "ASA") 

hub_map <- flights |>
  mutate(airline = str_sub(callsign, 1, 3)) |>
  filter(airline %in% target_airlines) |>
  count(airline, hub = departure, name = "flight_volume") |>
  group_by(airline) |>
  slice_max(order_by = flight_volume, n = 12, with_ties = FALSE) |>
  ungroup() |>
  dplyr::select(airline, hub)

own_hubs_tbl <- hub_map |>
  distinct(airline, dest = hub) |>
  mutate(is_own_hub = TRUE)

own_hubs_tbl
```

The top 12 airports for each airlines were selected as hubs. For airlines (such as United) that did not have this many airports, the future results of the code were not affected by this overestimate.

```{r add-regionals-to-mainlines, warning = FALSE}

regional_flights <- flights |>
  filter(str_detect(callsign, "^(RPA|SKW)")) |>
  pivot_longer(
    c(departure, arrival),
    names_to  = "leg",
    values_to = "airport"
  ) |>
  inner_join(hub_map, by = c("airport" = "hub")) |>
  count(icao24, airline, name = "hub_hits") |>
  group_by(icao24) |>
  slice_max(hub_hits, n = 1, with_ties = FALSE) |>
  ungroup()

flights <- flights |>
  left_join(regional_flights, by = join_by(icao24)) |>
  mutate(callsign = case_when(
    !is.na(airline) ~ str_c(airline, str_sub(callsign, 4)),
    str_sub(callsign, 1, 3) == "EDV" ~ str_c("DAL", str_sub(callsign, 4)),
    str_sub(callsign, 1, 3) == "QXE" ~ str_c("ASA", str_sub(callsign, 4)),
    str_sub(callsign, 1, 3) %in% c("JIA", "ENY", "PDT") ~ str_c("AAL", str_sub(callsign, 4)),
    str_sub(callsign, 1, 3) %in% c("ASH", "UCA", "GJS") ~ str_c("UAL", str_sub(callsign, 4)),
    TRUE ~ callsign
  ),
  airline = str_sub(callsign, 1, 3)) |>
  dplyr::select(-hub_hits)
```

The regional airlines are a huge part of the networks that are being flown. Some of the regional airlines only fly for a single larger counterpart (Endevour for Delta or Horizon for Alaska), but the larger Republic and Skywest fly for multiple airlines. Therefore, the number of hubs that each airplane takes-off or lands at was used to determine what marketable airline each one belonged to.

```{r}

valid_hubs <- hub_map |>
  distinct(airline, hub)

dep_counts <- flights |>
  mutate(
    airline = str_sub(callsign, 1, 3),
    dep_local = with_tz(firstseen, departure_tz),
    dep_hour = hour(dep_local)
  ) |>
  count(airline, hub = departure, dep_hour, name = "y") |>
  inner_join(valid_hubs, by = c("airline", "hub")) |>
  complete(nesting(airline, hub), dep_hour = 0:23, fill = list(y = 0))

hub_curves <- dep_counts |>
  group_by(airline, hub) |>
  group_map(~ {
    fit <- gam(
      y ~ s(dep_hour, bs = "cs", k = 10),
      family = poisson(),
      data = .x,
      method = "REML"
    )

    tibble(
      airline = .y$airline,
      hub = .y$hub,
      dep_hour = 0:23,
      lambda = predict(
        fit,
        newdata = tibble(dep_hour = 0:23),
        type = "response"
      )
    )
  }) |>
  bind_rows()

banks_from_curve <- function(df, win = 1) {
  df |>
    arrange(dep_hour) |>
    mutate(
      lam_s = slide_dbl(lambda, mean, .before = win, .after = win, .complete = TRUE),
      prev = lag(lam_s),
      next_val = lead(lam_s)
    ) |>
    mutate(is_valley = !is.na(prev) & !is.na(next_val) & lam_s < prev & lam_s < next_val) |>
    filter(is_valley) |>
    transmute(cut = dep_hour)
}

cuts <- hub_curves |>
  group_by(airline, hub) |>
  group_modify(~ banks_from_curve(.x)) |>
  ungroup()

bank_intervals <- cuts |>
  group_by(airline, hub) |>
  summarise(
    cuts = list(sort(unique(c(0, cut, 24)))),
    .groups = "drop"
  ) |>
  mutate(
    intervals = purrr::map(cuts, function(v) {
      tibble(
        start_hour = v[-length(v)],
        end_hour = v[-1],
        bank_idx = seq_along(v[-1])
      )
    })
  ) |>
  dplyr::select(-cuts) |>
  tidyr::unnest(intervals) |>
  mutate(bank = paste0("Bank ", bank_idx))

bank_intervals
```

```{r}

op_start <- 5
op_end <- 23

banks_from_curve_topk_hours <- function(curve_df, op_start = 6, op_end = 22, win = 1, min_banks = 3, max_banks = 4) {

  x <- curve_df |>
    arrange(hour) |>
    mutate(
      lam_s = slider::slide_dbl(lambda, mean, .before = win, .after = win, .complete = TRUE),
      prev = lag(lam_s),
      nextv = lead(lam_s),
      valley = !is.na(prev) & !is.na(nextv) & lam_s < prev & lam_s < nextv,
      depth = ((prev + nextv) / 2) - lam_s
    )

  valleys <- x |>
    filter(valley) |>
    arrange(desc(depth))

  n_cuts <- case_when(
    nrow(valleys) >= (max_banks - 1) ~ (max_banks - 1),
    nrow(valleys) >= (min_banks - 1) ~ (min_banks - 1),
    TRUE ~ NA_integer_
  )

  if (is.na(n_cuts)) {
    cuts <- seq(op_start, op_end, length.out = min_banks + 1)
  } 
  
  else {
    cuts_mid <- valleys |>
      slice_head(n = n_cuts) |>
      pull(hour)

    cuts <- sort(unique(c(op_start, cuts_mid, op_end)))

    if (length(cuts) < (min_banks + 1)) {
      cuts <- seq(op_start, op_end, length.out = min_banks + 1)
    }
  }

  tibble::tibble(
    start_hour = cuts[-length(cuts)],
    end_hour = cuts[-1],
    bank_idx = seq_along(cuts[-1]),
    bank = paste0("Bank ", bank_idx),
    start_min = start_hour * 60,
    end_min = end_hour * 60,
    n_banks = length(cuts) - 1
  )
}

bank_intervals <- hub_curves |>
  transmute(airline, hub, hour = dep_hour, lambda) |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y)
    banks_from_curve_topk_hours(
      .x,
      op_start = op_start,
      op_end = op_end,
      win = 1,
      min_banks = 3,
      max_banks = 4
    )
  ) |>
  ungroup()
```

```{r}

valid_hubs <- hub_map |>
  distinct(airline, hub)

dep_counts <- flights |>
  mutate(
    airline  = str_sub(callsign, 1, 3),
    dep_local = lubridate::with_tz(firstseen, departure_tz),
    dep_hour = lubridate::hour(dep_local)
  ) |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  filter(dep_hour >= op_start, dep_hour < op_end) |>
  count(airline, hub = departure, hour = dep_hour, name = "y") |>
  complete(nesting(airline, hub), hour = op_start:(op_end - 1), fill = list(y = 0))

arr_counts <- flights |>
  mutate(
    airline = str_sub(callsign, 1, 3),
    arr_local = lubridate::with_tz(lastseen, arrival_tz),
    arr_hour = lubridate::hour(arr_local)
  ) |>
  inner_join(valid_hubs, by = c("airline", "arrival" = "hub")) |>
  filter(arr_hour >= op_start, arr_hour < op_end) |>
  count(airline, hub = arrival, hour = arr_hour, name = "y") |>
  complete(nesting(airline, hub), hour = op_start:(op_end - 1), fill = list(y = 0))

fit_curve_window <- function(count_df) {
  fit <- mgcv::gam(
    y ~ s(hour, bs = "cs", k = 8),
    family = poisson(),
    data = count_df,
    method = "REML"
  )

  tibble(
    hour = sort(unique(count_df$hour)),
    lambda = predict(fit, newdata = tibble(hour = hour), type = "response")
  )
}

dep_curves <- dep_counts |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y) fit_curve_window(.x)) |>
  ungroup()

arr_curves <- arr_counts |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y) fit_curve_window(.x)) |>
  ungroup()

dep_bank_intervals <- dep_curves |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y) banks_from_curve_topk_hours(.x, win = 1, min_banks = 3, max_banks = 4)) |>
  ungroup()

arr_bank_intervals <- arr_curves |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y) banks_from_curve_topk_hours(.x, win = 1, min_banks = 3, max_banks = 4)) |>
  ungroup()
```

```{r}

tag_bank_snap <- function(df, intervals, minute_col, airport_col, bank_prefix) {

  join_by <- if (airport_col == "departure") {
    c("airline" = "airline", "departure" = "hub")
  } 
  
  else if (airport_col == "arrival") {
    c("airline" = "airline", "arrival" = "hub")
  } 
  
  else {
    stop("airport_col must be 'departure' or 'arrival'")
  }

  int_tbl <- intervals |>
    dplyr::select(airline, hub, bank, bank_idx, start_min, end_min) |>
    arrange(airline, hub, start_min)

  df |>
    inner_join(int_tbl, by = join_by) |>
    group_by(across(all_of(c("icao24","firstseen","lastseen","departure","arrival", "airline", airport_col)))) |>
    mutate(
      m = .data[[minute_col]],
      dist = case_when(
        m >= start_min & m < end_min ~ 0,
        m < start_min ~ start_min - m,
        TRUE ~ m - (end_min - 1)
      )
    ) |>
    slice_min(dist, n = 1, with_ties = FALSE) |>
    ungroup() |>
    transmute(
      icao24, firstseen, lastseen, departure, arrival,
      !!paste0(bank_prefix, "_hub")      := .data[[airport_col]],
      !!paste0(bank_prefix, "_bank")     := bank,
      !!paste0(bank_prefix, "_bank_idx") := bank_idx
    )
}

key_cols <- c("icao24", "firstseen", "lastseen", "departure", "arrival")

dep_tags <- flights |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  tag_bank_snap(dep_bank_intervals, minute_col = "dep_minute", 
                airport_col = "departure", bank_prefix = "dep")

arr_tags <- flights |>
  inner_join(valid_hubs, by = c("airline", "arrival" = "hub")) |>
  tag_bank_snap(arr_bank_intervals, minute_col = "arr_minute", airport_col = "arrival", bank_prefix = "arr")

flights_banked <- flights |>
  left_join(dep_tags, by = key_cols) |>
  left_join(arr_tags, by = key_cols)
```

```{r}

mct_min <- 35
max_ct_min <- 240

min_turn <- 20
max_turn <- 180

two_leg <- flights_banked |>
  arrange(icao24, firstseen) |>
  group_by(icao24) |>
  mutate(
    next_dep = lead(departure),
    next_arr = lead(arrival),
    next_start = lead(firstseen),
    turn_min = as.numeric(difftime(next_start, lastseen, units = "mins"))
  ) |>
  ungroup() |>
  filter(
    arrival == next_dep,
    turn_min >= min_turn, turn_min <= max_turn
  ) |>
  mutate(
    hub1 = departure,
    spoke = arrival,
    hub2 = next_arr
  ) |>
  inner_join(valid_hubs, by = c("airline", "hub1" = "hub")) |>
  inner_join(valid_hubs, by = c("airline", "hub2" = "hub")) |>
  filter(hub1 != hub2) |>
  count(airline, hub1, spoke, hub2, name = "n") |>
  arrange(desc(n))

two_leg
```

```{r}

wave_carriers <- c("DAL","AAL","ASA","UAL")

valid_hubs <- hub_map |>
  filter(airline %in% wave_carriers) |>
  distinct(airline, hub)

flights_wave <- flights |>
  filter(airline %in% wave_carriers) |>
  mutate(
    dep_local = lubridate::with_tz(firstseen, departure_tz),
    arr_local = lubridate::with_tz(lastseen,  arrival_tz),
    dep_minute = lubridate::hour(dep_local) * 60 + lubridate::minute(dep_local),
    arr_minute = lubridate::hour(arr_local) * 60 + lubridate::minute(arr_local),
    dep_hour = lubridate::hour(dep_local),
    arr_hour = lubridate::hour(arr_local),
    dep_date = as.Date(dep_local),
    arr_date = as.Date(arr_local)
  )
```

```{r}

fit_curve <- function(count_df) {
  fit <- mgcv::gam(
    y ~ s(hour, bs = "cs", k = 10),
    family = poisson(),
    data   = count_df,
    method = "REML"
  )
  tibble(hour = 0:23, lambda = predict(fit, newdata = tibble(hour = 0:23), type = "response"))
}

dep_counts <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  count(airline, hub = departure, hour = dep_hour, name = "y") |>
  complete(nesting(airline, hub), hour = 0:23, fill = list(y = 0))

arr_counts <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "arrival" = "hub")) |>
  count(airline, hub = arrival, hour = arr_hour, name = "y") |>
  complete(nesting(airline, hub), hour = 0:23, fill = list(y = 0))

dep_banks <- dep_counts |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y) {
    curve <- fit_curve(.x)
    banks_from_curve_topk_hours(curve, win = 1, min_banks = 3, max_banks = 4)
  }) |>
  ungroup()

arr_banks <- arr_counts |>
  group_by(airline, hub) |>
  group_modify(\(.x, .y) {
    curve <- fit_curve(.x)
    banks_from_curve_topk_hours(curve, win = 1, min_banks = 3, max_banks = 4)
  }) |>
  ungroup()
```

```{r}

connect_score <- function(t, arr_vec, mct = 35, max_ct = 180) {
  t <- as.integer(t) %% 1440
  lo <- t - max_ct
  hi <- t - mct

  rng <- seq(lo, hi)
  rng_mod <- (rng %% 1440) + 1
  sum(arr_vec[rng_mod])
}

optimize_slots_in_bank <- function(start_min, end_min, n_slots, arr_vec,
                                   mct = 35, max_ct = 180,
                                   buffer = 5, min_sep = 4,
                                   search_radius = 30, passes = 2) {
  if (n_slots <= 0) return(integer())
  lo <- start_min + buffer
  hi <- end_min   - buffer
  if (hi <= lo) return(rep(round((start_min + end_min) / 2), n_slots))

  slots <- if (n_slots == 1) {
    round((lo + hi) / 2)
  } else {
    round(seq(lo, hi, length.out = n_slots))
  }

  for (p in seq_len(passes)) {
    for (i in seq_along(slots)) {
      left_bound  <- if (i == 1) lo else slots[i - 1] + min_sep
      right_bound <- if (i == length(slots)) hi else slots[i + 1] - min_sep
      if (right_bound < left_bound) next

      cand_lo <- max(left_bound,  slots[i] - search_radius)
      cand_hi <- min(right_bound, slots[i] + search_radius)
      cands <- cand_lo:cand_hi

      scores <- vapply(
        cands,
        \(tt) connect_score(tt, arr_vec, mct = mct, max_ct = max_ct),
        numeric(1)
      )

      slots[i] <- cands[which.max(scores)]
    }
  }

  slots
}
```

```{r}

hub_dep_obs <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  left_join(dep_banks |> dplyr::select(airline, hub, bank, bank_idx, start_min, end_min),
            by = c("airline", "departure" = "hub")) |>
  filter(!is.na(start_min), !is.na(end_min)) |>
  filter(dep_minute >= start_min, dep_minute < end_min) |>
  mutate(hub = departure) |>
  group_by(airline, hub, bank, bank_idx, start_min, end_min, dep_date) |>
  summarise(n_day = n(), .groups = "drop") |>
  group_by(airline, hub, bank, bank_idx, start_min, end_min) |>
  complete(
    dep_date = seq(min(dep_date), max(dep_date), by = "day"),
    fill = list(n_day = 0)
  ) |>
  summarise(
    n_dep = as.integer(ceiling(stats::quantile(n_day, 0.90, na.rm = TRUE))),
    .groups = "drop"
  ) |>
  mutate(n_dep = pmax(0L, n_dep))

route_shares <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  left_join(
    dep_banks |> dplyr::select(airline, hub, bank, bank_idx, start_min, end_min),
    by = c("airline", "departure" = "hub")
  ) |>
  filter(dep_minute >= start_min, dep_minute < end_min) |>
  mutate(hub = departure, dest = arrival) |>
  count(airline, hub, bank, bank_idx, dest, name = "n") |>
  left_join(own_hubs_tbl, by = c("airline", "dest")) |>
  mutate(
    is_own_hub = coalesce(is_own_hub, FALSE),
    w = if_else(is_own_hub, 0.5, 1.0),   # <-- this is your “split in half”
    n_w = n * w
  ) |>
  group_by(airline, hub, bank, bank_idx) |>
  mutate(p = n_w / sum(n_w)) |>
  ungroup()

route_dist <- route_shares |>
  group_by(airline, hub, bank, bank_idx) |>
  summarise(dests = list(dest), probs = list(p), .groups = "drop")

arr_vec_tbl <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "arrival" = "hub")) |>
  mutate(hub = arrival) |>
  group_by(airline, hub) |>
  summarise(
    n_days = n_distinct(arr_date),
    arr_vec = list(
      tabulate((arr_minute %% 1440) + 1L, nbins = 1440) / pmax(1L, n_days)
    ),
    .groups = "drop"
  )

set.seed(1)

mct_min   <- 35
max_ct_min <- 180

allocate_dests_bank <- function(n_slots, dests, probs) {
  if (n_slots <= 0) return(character())
  if (is.null(dests) || length(dests) == 0) return(rep(NA_character_, n_slots))

  # 1. Guarantee coverage: Create a list containing ONE of every destination
  # available in the data, regardless of probability.
  guaranteed_deck <- dests
  
  # 2. Weighted fill: If we need more slots than unique destinations, 
  # fill the rest using the probabilities (popular routes get more flights).
  if (n_slots > length(guaranteed_deck)) {
    remainder <- n_slots - length(guaranteed_deck)
    
    # normalize probs
    probs <- as.numeric(probs)
    probs[is.na(probs) | probs < 0] <- 0
    if(sum(probs) > 0) probs <- probs / sum(probs)
    
    extra_flights <- sample(dests, size = remainder, replace = TRUE, prob = probs)
    final_list <- c(guaranteed_deck, extra_flights)
    
  } else {
    # If we have fewer slots than destinations, we must sample without replacement
    # but we prioritize by probability to keep the major routes alive.
    final_list <- sample(dests, size = n_slots, replace = FALSE, prob = probs)
  }

  # Shuffle so the small destinations aren't always at the end of the bank
  sample(final_list)
}

set.seed(1)

wave_schedule <- hub_dep_obs |>
  left_join(arr_vec_tbl, by = c("airline", "hub")) |>
  mutate(
    slot_minute = purrr::pmap(
      list(start_min, end_min, n_dep, arr_vec),
      \(.s, .e, .n, .v) optimize_slots_in_bank(
        .s, .e, .n, .v,
        mct = mct_min, max_ct = max_ct_min,
        buffer = 5, min_sep = 4, search_radius = 30, passes = 2
      )
    )
  ) |>
  tidyr::unnest(slot_minute) |>
  mutate(slot_minute = as.integer(round(slot_minute))) |>
  group_by(airline, hub, bank, bank_idx) |>
  mutate(
    slot_idx = row_number(),
    connect  = vapply(
      slot_minute,
      \(tt) connect_score(tt, arr_vec[[1]], mct = mct_min, max_ct = max_ct_min),
      numeric(1)
    )
  ) |>
  ungroup() |>
  left_join(route_dist, by = c("airline","hub","bank","bank_idx")) |>
  group_by(airline, hub, bank, bank_idx) |>
  mutate(
    dest_vec = list(allocate_dests_bank(first(n_dep), dests[[1]], probs[[1]])),
    dest     = dest_vec[[1]][slot_idx],
    dep_hhmm = sprintf(
      "%02d:%02d",
      as.integer(slot_minute %/% 60),
      as.integer(slot_minute %% 60)
    )
  ) |>
  ungroup() |>
  dplyr::select(-dests, -probs, -dest_vec) |>
  arrange(airline, hub, bank_idx, slot_minute)
```

```{r}

route_block <- flights_wave |>
  mutate(dur_sec = as.numeric(difftime(lastseen, firstseen, units = "secs"))) |>
  filter(is.finite(dur_sec), dur_sec > 10 * 60, dur_sec < 10 * 3600) |>
  group_by(airline, departure, arrival) |>
  summarise(
    n = n(),
    dur_sec_med = median(dur_sec, na.rm = TRUE),
    .groups = "drop"
  )

get_block_sec <- function(airline, dep, arr, route_block, fallback = 90 * 60) {
  hit <- route_block |>
    filter(airline == !!airline, departure == !!dep, arrival == !!arr)
  if (nrow(hit) == 1 && hit$n >= 10 && is.finite(hit$dur_sec_med)) hit$dur_sec_med else fallback
}
```

```{r}

get_duration_safe <- function(al, dep, arr, r_block) {
  match_row <- r_block |> 
    filter(airline == al, departure == dep, arrival == arr)
  
  if(nrow(match_row) > 0) return(as.integer(match_row$dur_sec_med))
  
  return(as.integer(3600)) 
}

add_coords_11cols <- function(tbl, airports_ref) {
  dep_coords <- airports_ref |> 
    dplyr::select(ICAO, dep_latitude = lat, dep_longitude = lon) |> 
    distinct()
  
  arr_coords <- airports_ref |> 
    dplyr::select(ICAO, arr_latitude = lat, arr_longitude = lon) |> 
    distinct()
  
  tbl |> 
    left_join(dep_coords, by = c("departure" = "ICAO")) |> 
    left_join(arr_coords, by = c("arrival" = "ICAO")) |> 
    dplyr::select(
      any_of(c("airline", "callsign", "tail_number", "departure", "arrival", 
               "day_of_week", "departure_time", "arrival_time")),
      dep_latitude, dep_longitude, arr_latitude, arr_longitude
    )
}

base_demand <- wave_schedule |>
  mutate(
    dep_sec = as.integer(round(slot_minute)) * 60L,
    airline_code = airline
  ) |>
  rowwise() |>
  mutate(
    duration_out = get_duration_safe(airline, hub, dest, route_block)
  ) |>
  ungroup() |>
  transmute(
    airline,
    flight_id_base = paste0(airline, "-", hub, "-", bank_idx, "-", row_number()),
    origin = hub,
    dest = dest,
    dep_time = dep_sec,
    arr_time = dep_sec + duration_out
  )

min_turn_seconds <- 45 * 60 

return_demand <- base_demand |>
    rowwise() |>
    mutate(
      # Turn time is between 45 mins and 3 hours (randomized)
      actual_turn = as.integer(runif(1, 45 * 60, 180 * 60)),
      
      origin_ret = dest, 
      dest_ret = origin,
      
      # Departure is Arrival + Turn
      dep_time_ret = arr_time + actual_turn
    ) |>
  rowwise() |>
  mutate(
    duration_in = get_duration_safe(airline, origin_ret, dest_ret, route_block)
  ) |>
  ungroup() |>
  transmute(
    airline,
    flight_id_base = paste0(flight_id_base, "R"),
    origin = origin_ret,
    dest = dest_ret,
    dep_time = dep_time_ret,
    arr_time = dep_time_ret + duration_in
  )

master_schedule <- bind_rows(
  base_demand |> mutate(type = "outbound"),
  return_demand |> mutate(type = "inbound")
) |>
  arrange(dep_time)

fleet_state <- tibble(
  airline = character(),
  tail_num = character(),
  loc = character(),
  ready_at = integer()
)

assigned_schedule <- list()
tail_counter <- 1 
turn_buffer_hub <- 60 * 60

total_flights <- nrow(master_schedule)
cat("Assigning aircraft to", total_flights, "flights...\n")
pb <- txtProgressBar(min = 0, max = total_flights, style = 3)

for(i in 1:total_flights) {
  
  flight <- master_schedule[i, ]

  candidates <- fleet_state |>
    filter(
      airline == flight$airline,
      loc == flight$origin,
      ready_at <= flight$dep_time
    ) |>
    arrange(desc(ready_at))
  
  if(nrow(candidates) > 0) {
    chosen_tail <- candidates$tail_num[1]
    
    fleet_state <- fleet_state |>
      filter(tail_num != chosen_tail) |>
      add_row(
        airline = flight$airline,
        tail_num = chosen_tail,
        loc = flight$dest,
        ready_at = flight$arr_time + turn_buffer_hub
      )
    
  } else {
    chosen_tail <- paste0("N", str_pad(tail_counter, 5, pad = "0"), flight$airline)
    tail_counter <- tail_counter + 1
    
    fleet_state <- fleet_state |>
      add_row(
        airline = flight$airline,
        tail_num = chosen_tail,
        loc = flight$dest,
        ready_at = flight$arr_time + turn_buffer_hub
      )
  }
  
  assigned_schedule[[i]] <- flight |> mutate(tail_number = chosen_tail)
  
  setTxtProgressBar(pb, i)
}
close(pb)

final_daily_schedule <- bind_rows(assigned_schedule)

airports_ref <- airportr::airports |> 
  dplyr::select(ICAO, lat = Latitude, lon = Longitude) |> 
  distinct()

airport_tz_map <- airports_ref |> 
  filter(!is.na(lat), !is.na(lon)) |> 
  dplyr::select(ICAO, lat, lon) |> 
  mutate(tz_name = lutz::tz_lookup_coords(lat, lon, method = "fast")) |> 
  dplyr::select(ICAO, tz_name)

ref_date_str <- as.character(Sys.Date()) 

schedule_et <- final_daily_schedule |> 
  left_join(airport_tz_map, by = c("origin" = "ICAO")) |> 
  filter(!is.na(tz_name)) |> 
  group_by(tz_name) |> 
  mutate(

    midnight_local = lubridate::ymd_hms(paste(ref_date_str, "00:00:00"), tz = tz_name[1]),
    
    local_dep_dt = midnight_local + seconds(dep_time),
    
    et_dep_dt = with_tz(local_dep_dt, "America/New_York"),
    
    flight_duration = arr_time - dep_time,
    et_arr_dt = et_dep_dt + seconds(flight_duration)
  ) |> 
  ungroup() |> 
  mutate(
    dep_time_hms = as_hms(et_dep_dt),
    arr_time_hms = as_hms(et_arr_dt),
    day_offset = as.integer(floor(difftime(floor_date(et_dep_dt, "day"), 
                                           as.POSIXct(ref_date_str, tz="America/New_York"), 
                                           units = "days")))
  )

day_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

full_week_schedule <- map_dfr(1:7, function(i) {
  
  schedule_et |> 
    mutate(
      current_day_idx = i,
      final_day_idx = (current_day_idx + day_offset - 1) %% 7 + 1,
      day_of_week = day_levels[final_day_idx],
      callsign = paste0(airline, str_sub(tail_number, 2, 5))
    )
}) |> 
  rename(
    departure = origin,
    arrival = dest,
    departure_time = dep_time_hms,
    arrival_time = arr_time_hms
  ) |> 
  add_coords_11cols(airports_ref = airports_ref) |> 
  dplyr::select(
    airline, callsign, tail_number, departure, arrival, 
    day_of_week, departure_time, arrival_time, 
    dep_latitude, dep_longitude, arr_latitude, arr_longitude
  ) |> 
  mutate(
    duration = as.numeric(arrival_time - departure_time), 
    day_of_week = factor(day_of_week, levels = day_levels, ordered = TRUE)
  ) |> 
  arrange(day_of_week, departure_time)

# --- 1. PREPARE DAILY DEMAND & PROBABILITIES ---

# Calculate Departure Volume per Weekday (Mon, Tue, etc.)
hub_dep_obs_daily <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  left_join(dep_banks |> dplyr::select(airline, hub, bank, bank_idx, start_min, end_min),
            by = c("airline", "departure" = "hub")) |>
  filter(!is.na(start_min), !is.na(end_min)) |>
  filter(dep_minute >= start_min, dep_minute < end_min) |>
  mutate(
    hub = departure,
    # Extract Day Name (Monday, Tuesday...)
    day_name = lubridate::wday(dep_date, label = TRUE, abbr = FALSE) 
  ) |>
  group_by(airline, hub, bank, bank_idx, start_min, end_min, day_name, dep_date) |>
  summarise(n_day = n(), .groups = "drop") |>
  group_by(airline, hub, bank, bank_idx, start_min, end_min, day_name) |>
  summarise(
    # Use 90th percentile to ensure we capture "busy" days properly
    n_dep = as.integer(ceiling(stats::quantile(n_day, 0.90, na.rm = TRUE))),
    .groups = "drop"
  ) |>
  mutate(n_dep = pmax(0L, n_dep))

# Calculate Route Probabilities (Global - aggregated across all days for stability)
route_shares <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "departure" = "hub")) |>
  left_join(
    dep_banks |> dplyr::select(airline, hub, bank, bank_idx, start_min, end_min),
    by = c("airline", "departure" = "hub")
  ) |>
  filter(dep_minute >= start_min, dep_minute < end_min) |>
  mutate(hub = departure, dest = arrival) |>
  count(airline, hub, bank, bank_idx, dest, name = "n") |>
  left_join(own_hubs_tbl, by = c("airline", "dest")) |>
  mutate(
    is_own_hub = coalesce(is_own_hub, FALSE),
    w = if_else(is_own_hub, 0.5, 1.0),
    n_w = n * w
  ) |>
  group_by(airline, hub, bank, bank_idx) |>
  mutate(p = n_w / sum(n_w)) |>
  ungroup()

route_dist <- route_shares |>
  group_by(airline, hub, bank, bank_idx) |>
  summarise(dests = list(dest), probs = list(p), .groups = "drop")

# Calculate Arrival Profiles (Global)
arr_vec_tbl <- flights_wave |>
  inner_join(valid_hubs, by = c("airline", "arrival" = "hub")) |>
  mutate(hub = arrival) |>
  group_by(airline, hub) |>
  summarise(
    n_days = n_distinct(arr_date),
    arr_vec = list(
      tabulate((arr_minute %% 1440) + 1L, nbins = 1440) / pmax(1L, n_days)
    ),
    .groups = "drop"
  )

# --- 2. HELPER FUNCTIONS ---

set.seed(1)
mct_min    <- 35
max_ct_min <- 180

allocate_dests_bank <- function(n_slots, dests, probs) {
  if (n_slots <= 0) return(character())
  if (is.null(dests) || length(dests) == 0) return(rep(NA_character_, n_slots))

  probs <- as.numeric(probs)
  probs[is.na(probs) | probs < 0] <- 0
  if (sum(probs) <= 0) return(rep(NA_character_, n_slots))
  probs <- probs / sum(probs)

  exp_ct <- n_slots * probs
  base   <- floor(exp_ct)
  rem    <- n_slots - sum(base)
  frac   <- exp_ct - base

  if (rem > 0) {
    add_idx <- order(frac, decreasing = TRUE)[seq_len(rem)]
    base[add_idx] <- base[add_idx] + 1L
  }

  vec <- rep(dests, base)
  sample(vec, size = n_slots, replace = FALSE)
}

get_duration_safe <- function(al, dep, arr, r_block) {
  match_row <- r_block |> 
    filter(airline == al, departure == dep, arrival == arr)
  
  if(nrow(match_row) > 0) return(as.integer(match_row$dur_sec_med))
  return(as.integer(3600)) 
}

# Core function to generate one day of flights (unassigned tails)
generate_daily_schedule <- function(target_day, obs_data, routes_data, arr_vec_data) {
  
  # Filter demand for the specific day (e.g., "Monday")
  day_obs <- obs_data |> filter(day_name == target_day)
  
  if(nrow(day_obs) == 0) return(tibble())

  # Optimize slots
  day_schedule <- day_obs |>
    left_join(arr_vec_data, by = c("airline", "hub")) |>
    mutate(
      slot_minute = purrr::pmap(
        list(start_min, end_min, n_dep, arr_vec),
        \(.s, .e, .n, .v) optimize_slots_in_bank(
          .s, .e, .n, .v,
          mct = mct_min, max_ct = max_ct_min,
          buffer = 5, min_sep = 4, search_radius = 30, passes = 2
        )
      )
    ) |>
    tidyr::unnest(slot_minute) |>
    mutate(slot_minute = as.integer(round(slot_minute))) |>
    
    # Check connectivity scores (optional but good for debugging)
    group_by(airline, hub, bank, bank_idx) |>
    mutate(
      slot_idx = row_number(),
      connect = vapply(slot_minute, \(tt) connect_score(tt, arr_vec[[1]]), numeric(1))
    ) |>
    ungroup() |>
    
    # Assign destinations
    left_join(routes_data, by = c("airline","hub","bank","bank_idx")) |>
    group_by(airline, hub, bank, bank_idx) |>
    mutate(
      dest_vec = list(allocate_dests_bank(first(n_dep), dests[[1]], probs[[1]])),
      dest     = dest_vec[[1]][slot_idx]
    ) |>
    ungroup() |>
    dplyr::select(-dests, -probs, -dest_vec) |>
    filter(!is.na(dest)) 

  # Calculate block times and returns
  base_demand <- day_schedule |>
    mutate(dep_sec = as.integer(round(slot_minute)) * 60L) |>
    rowwise() |>
    mutate(duration_out = get_duration_safe(airline, hub, dest, route_block)) |>
    ungroup() |>
    transmute(
      airline,
      flight_id_base = paste0(airline, "-", hub, "-", bank_idx, "-", row_number()),
      origin = hub, dest = dest,
      dep_time = dep_sec,
      arr_time = dep_sec + duration_out
    )

  min_turn_seconds <- 45 * 60 

  return_demand <- base_demand |>
    mutate(
      origin_ret = dest, dest_ret = origin,
      dep_time_ret = arr_time + min_turn_seconds
    ) |>
    rowwise() |>
    mutate(duration_in = get_duration_safe(airline, origin_ret, dest_ret, route_block)) |>
    ungroup() |>
    transmute(
      airline,
      flight_id_base = paste0(flight_id_base, "R"),
      origin = origin_ret, dest = dest_ret,
      dep_time = dep_time_ret,
      arr_time = dep_time_ret + duration_in
    )

  # Combine outbound and inbound
  bind_rows(base_demand, return_demand) |>
    mutate(day_of_week = target_day) |>
    arrange(dep_time)
}

# --- 3. RUN 7-DAY SIMULATION ---

# Initialize empty fleet
fleet_state <- tibble(
  airline = character(),
  tail_num = character(),
  loc = character(),
  ready_at = integer() # seconds from start of the *current* day
)

tail_counter <- 1
turn_buffer_hub <- 45 * 60
seconds_per_day <- 24 * 60 * 60

days_to_sim <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
full_week_results <- list()

cat("Starting 7-day continuous simulation...\n")

for (d_idx in seq_along(days_to_sim)) {
  current_day <- days_to_sim[d_idx]
  cat("Simulating:", current_day, "\n")
  
  # A. Generate demand specific to this day of the week
  daily_flights <- generate_daily_schedule(current_day, hub_dep_obs_daily, route_dist, arr_vec_tbl)
  
  # B. Reset fleet clocks for the new day
  # If a plane was ready at 26:00 (2 AM) on Monday, it is ready at 02:00 on Tuesday
  if (nrow(fleet_state) > 0) {
    fleet_state <- fleet_state |>
      mutate(ready_at = pmax(0, ready_at - seconds_per_day))
  }
  
  # C. Assign tails
  assigned_day <- list()
  daily_flights <- daily_flights |> arrange(dep_time)
  
  if(nrow(daily_flights) > 0) {
    for(i in seq_len(nrow(daily_flights))) {
      flight <- daily_flights[i, ]
      
      # Look for existing aircraft
      candidates <- fleet_state |>
        filter(
          airline == flight$airline,
          loc == flight$origin,
          ready_at <= flight$dep_time
        ) |>
        arrange(desc(ready_at)) 
      
      if(nrow(candidates) > 0) {
        chosen_tail <- candidates$tail_num[1]
        
        fleet_state <- fleet_state |>
          filter(tail_num != chosen_tail) |>
          add_row(
            airline  = flight$airline,
            tail_num = chosen_tail,
            loc      = flight$dest,
            ready_at = flight$arr_time + turn_buffer_hub
          )
      } else {
        # Create new aircraft
        chosen_tail <- paste0("N", stringr::str_pad(tail_counter, 5, pad = "0"), flight$airline)
        tail_counter <- tail_counter + 1
        
        fleet_state <- fleet_state |>
          add_row(
            airline  = flight$airline,
            tail_num = chosen_tail,
            loc      = flight$dest,
            ready_at = flight$arr_time + turn_buffer_hub
          )
      }
      
      assigned_day[[i]] <- flight |> mutate(tail_number = chosen_tail)
    }
    full_week_results[[d_idx]] <- bind_rows(assigned_day)
  }
}

final_week_schedule <- bind_rows(full_week_results)

# --- 4. FORMATTING AND SAVING ---

add_coords_11cols <- function(tbl, airports_ref) {
  dep_coords <- airports_ref |> 
    dplyr::select(ICAO, dep_latitude = lat, dep_longitude = lon) |> 
    distinct()
  
  arr_coords <- airports_ref |> 
    dplyr::select(ICAO, arr_latitude = lat, arr_longitude = lon) |> 
    distinct()
  
  tbl |> 
    left_join(dep_coords, by = c("departure" = "ICAO")) |> 
    left_join(arr_coords, by = c("arrival" = "ICAO")) |> 
    dplyr::select(
      any_of(c("airline", "callsign", "tail_number", "departure", "arrival", 
               "day_of_week", "departure_time", "arrival_time")),
      dep_latitude, dep_longitude, arr_latitude, arr_longitude
    )
}

airports_ref <- airportr::airports |> 
  dplyr::select(ICAO, lat = Latitude, lon = Longitude) |> 
  distinct()

final_formatted <- final_week_schedule |>
  rename(departure = origin, arrival = dest) |>
  mutate(
    callsign = paste0(airline, str_sub(tail_number, 2, 5)),
    departure_time = hms::as_hms(dep_time),
    arrival_time = hms::as_hms(arr_time),
    day_of_week = factor(day_of_week, levels = days_to_sim, ordered = TRUE)
  ) |>
  add_coords_11cols(airports_ref = airports_ref) |>
  mutate(duration = as.numeric(arrival_time - departure_time)) |> 
  arrange(day_of_week, departure_time)

saveRDS(final_formatted, "hub_spoke_schedule.rds")
```

### Point-to-Point Carriers

```{r setting-up-point-to-point}

ptp_prefix <- c("FFT", "JBU", "NKS", "SWA")

flights <- all_data |>
  mutate(
    airline = stringr::str_sub(callsign, 1, 3),
    firstseen_utc = lubridate::ymd_hms(firstseen, tz = "UTC"),
    lastseen_utc = lubridate::ymd_hms(lastseen, tz = "UTC"),
    service_date_utc = as.Date(firstseen_utc)
  ) |>
  filter(
    departure %in% us_airports$ICAO,
    arrival %in% us_airports$ICAO,
    departure != arrival,
    airline %in% ptp_prefix
    ) |>
  mutate(
    duration_s = as.numeric(difftime(lastseen_utc, firstseen_utc, units = "secs")),
    duration_m = duration_s / 60
    ) |>
  filter(is.finite(duration_m), duration_m > 0, duration_m < 24 * 60) |>
  dplyr::select(
    airline, callsign, icao24, departure, arrival, firstseen_utc, lastseen_utc, 
    service_date_utc, duration_m
  )

spread_midnight_spike_utc <- function(
  flights_tbl,
  airlines = c("SWA","FFT","NKS","JBU"),
  max_back_hours = 2L,
  expected_method = c("p50_1to3", "hour1"),
  seed = 1
) {
  expected_method <- match.arg(expected_method)
  set.seed(seed)

  x <- flights_tbl |>
    mutate(
      airline = as.character(airline),
      firstseen_utc = as.POSIXct(firstseen_utc, tz = "UTC"),
      lastseen_utc  = as.POSIXct(lastseen_utc,  tz = "UTC"),
      service_date_utc = as.Date(firstseen_utc),
      hour_utc = lubridate::hour(firstseen_utc),
      .row_id = row_number()
    )

  hourly <- x |>
    filter(airline %in% airlines) |>
    count(airline, service_date_utc, hour_utc, name = "n") |>
    tidyr::complete(airline, service_date_utc, hour_utc = 0:23, fill = list(n = 0))

  expected <- hourly |>
    group_by(airline, service_date_utc) |>
    summarise(
      n0 = n[hour_utc == 0],
      n1 = n[hour_utc == 1],
      n2 = n[hour_utc == 2],
      n3 = n[hour_utc == 3],
      expected0 = dplyr::case_when(
        expected_method == "hour1"    ~ n1,
        expected_method == "p50_1to3" ~ as.integer(stats::median(c(n1, n2, n3))),
        TRUE ~ n1
      ),
      excess0 = pmax(0L, n0 - expected0),
      .groups = "drop"
    ) |>
    filter(excess0 > 0)

  if (nrow(expected) == 0) {
    return(x |> dplyr::select(-hour_utc, -.row_id))
  }

  cand <- x |>
    filter(airline %in% airlines, hour_utc == 0) |>
    inner_join(expected |> dplyr::select(airline, service_date_utc, excess0),
               by = c("airline", "service_date_utc"))

  if (nrow(cand) == 0) {
    return(x |> dplyr::select(-hour_utc, -.row_id))
  }

  movers <- cand |>
    group_by(airline, service_date_utc) |>
    group_modify(\(.df, .key) {
      n_move <- min(nrow(.df), .df$excess0[[1]])
      if (n_move <= 0) return(.df[0, ])
      dplyr::slice_sample(.df, n = n_move)
    }) |>
    ungroup() |>
    dplyr::select(.row_id)

  if (nrow(movers) == 0) {
    return(x |> dplyr::select(-hour_utc, -.row_id))
  }

  ks <- 1L:max_back_hours
  probs <- rev(seq_len(max_back_hours))
  probs <- probs / sum(probs)

  shift_tbl <- movers |>
    mutate(k = sample(ks, size = n(), replace = TRUE, prob = probs))

  out <- x |>
    left_join(shift_tbl, by = ".row_id") |>
    mutate(
      firstseen_utc = if_else(!is.na(k), firstseen_utc - lubridate::hours(k), firstseen_utc),
      lastseen_utc  = if_else(!is.na(k), lastseen_utc  - lubridate::hours(k), lastseen_utc),
      service_date_utc = as.Date(firstseen_utc)
    ) |>
    dplyr::select(-hour_utc, -.row_id, -k)

  out
}

flights <- spread_midnight_spike_utc(
  flights,
  airlines = ptp_prefix,
  max_back_hours = 2,
  expected_method = "p50_1to3",
  seed = 1
)

airports_ref <- airportr::airports |>
  transmute(
    ICAO = stringr::str_to_upper(stringr::str_trim(ICAO)),
    lat  = Latitude,
    lon  = Longitude
  ) |>
  distinct() |>
  filter(!is.na(lat), !is.na(lon))

tz_lookup <- airports_ref |>
  rowwise() |>
  mutate(timezone = lutz::tz_lookup_coords(lat, lon, method = "fast")) |>
  ungroup() |>
  filter(!is.na(timezone))

dep_tz_tbl <- tz_lookup |>
  dplyr::select(ICAO, departure_tz = timezone)

arr_tz_tbl <- tz_lookup |>
  dplyr::select(ICAO, arrival_tz = timezone)

flights <- flights |>
  left_join(dep_tz_tbl, by = c("departure" = "ICAO")) |>
  left_join(arr_tz_tbl, by = c("arrival" = "ICAO")) |>
  mutate(
    dep_local  = lubridate::with_tz(firstseen_utc, departure_tz),
    arr_local  = lubridate::with_tz(lastseen_utc,  arrival_tz),
    dep_minute_local = lubridate::hour(dep_local) * 60L + lubridate::minute(dep_local),
    arr_minute_local = lubridate::hour(arr_local) * 60L + lubridate::minute(arr_local)
  )

WINDOW_START <- 5 * 60
WINDOW_END <- 22 * 60

flights <- flights |>
  filter(dep_minute_local >= WINDOW_START, dep_minute_local <= WINDOW_END)

flights |>
  count(airline, service_date_utc, name = "n_flights") |>
  group_by(airline) |>
  summarise(
    days = n(),
    avg_flights = mean(n_flights),
    max_flights = max(n_flights),
    .groups = "drop"
  )
```

```{r}

clean_flights <- all_data |>
  mutate(
    airline = stringr::str_sub(callsign, 1, 3),
    firstseen_utc = lubridate::ymd_hms(firstseen, tz = "UTC"),
    lastseen_utc  = lubridate::ymd_hms(lastseen, tz = "UTC"),
    duration_min = as.numeric(difftime(lastseen_utc, firstseen_utc, units = "mins"))
  ) |>
  filter(
    departure %in% us_airports$ICAO,
    arrival %in% us_airports$ICAO,
    departure != arrival,
    airline %in% ptp_prefix,
    duration_min > 20, duration_min < 600
  )

fleet_sizes <- clean_flights |>
  group_by(airline) |>
  summarise(n_planes = n_distinct(icao24)) |>
  tibble::deframe()

route_block_times <- clean_flights |>
  group_by(airline, departure, arrival) |>
  summarise(
    block_min = as.integer(median(duration_min)),
    .groups = "drop"
  )

transition_matrix <- clean_flights |>
  count(airline, departure, arrival, name = "flights") |>
  group_by(airline, departure) |>
  mutate(prob = flights / sum(flights)) |>
  ungroup() |>
  dplyr::select(airline, origin = departure, dest = arrival, prob)

starting_weights <- clean_flights |>
  count(airline, departure, name = "weight") |>
  group_by(airline) |>
  mutate(prob = weight / sum(weight)) |>
  ungroup()

simulate_airline <- function(airline_code, n_planes, days = 7) {
  
  my_trans <- transition_matrix |> filter(airline == airline_code)
  my_blocks <- route_block_times |> filter(airline == airline_code)
  my_starts <- starting_weights |> filter(airline == airline_code)
  
  if(nrow(my_starts) == 0) return(NULL) # Skip if no data
  
  start_sim <- as.POSIXct("2026-01-01 05:00:00", tz = "UTC")
  end_sim   <- start_sim + days(days)
  
  fleet <- tibble(
    tail_number = sprintf("N%04d%s", 1:n_planes, airline_code),
    airline = airline_code,
    current_airport = sample(my_starts$departure, n_planes, replace = TRUE, prob = my_starts$prob),
    # Stagger starts between 05:00 and 09:00 UTC to avoid a massive "simultaneous takeoff" spike
    next_available = start_sim + minutes(sample(0:240, n_planes, replace = TRUE)) 
  )
  
  schedule_log <- list()
  
  cat(sprintf("Simulating %s with %d planes...\n", airline_code, n_planes))
  pb <- txtProgressBar(min = 0, max = n_planes, style = 3)
  
  for (p in 1:n_planes) {
    
    curr_tail <- fleet$tail_number[p]
    curr_loc  <- fleet$current_airport[p]
    curr_time <- fleet$next_available[p]
    
    plane_legs <- list()
    
    while (curr_time < end_sim) {
      
      routes <- my_trans |> filter(origin == curr_loc)
      
      if (nrow(routes) == 0) {
        curr_loc <- sample(my_starts$departure, 1, prob = my_starts$prob)
        curr_time <- curr_time + hours(4) 
        next
      }
      
      next_dest <- sample(routes$dest, 1, prob = routes$prob)
      
      blk <- my_blocks |> 
        filter(departure == curr_loc, arrival == next_dest) |> 
        pull(block_min)
      
      if(length(blk) == 0) blk <- 120
      
      dep_time <- curr_time
      arr_time <- dep_time + minutes(blk)
      
      utc_hour <- lubridate::hour(dep_time)
      if (utc_hour >= 6 && utc_hour < 10) { 
        dep_time <- dep_time + hours(6)
        arr_time <- dep_time + minutes(blk)
      }
      
      plane_legs[[length(plane_legs) + 1]] <- tibble(
        airline = airline_code,
        tail_number = curr_tail,
        departure = curr_loc,
        arrival = next_dest,
        dep_time = dep_time,
        arr_time = arr_time,
        duration = blk
      )
      
      curr_loc <- next_dest
      curr_time <- arr_time + minutes(45)
    }
    
    schedule_log[[p]] <- bind_rows(plane_legs)
    setTxtProgressBar(pb, p)
  }
  close(pb)
  
  bind_rows(schedule_log)
}

sim_results <- list()

for (al in names(fleet_sizes)) {
  sim_results[[al]] <- simulate_airline(al, fleet_sizes[[al]])
}

full_gen_schedule <- bind_rows(sim_results)

airport_locs_clean <- airport_lookup_1 |> 
  transmute(ICAO = ICAO, lat = latitude, lon = longitude) |> 
  filter(!is.na(lat), !is.na(lon)) |> 
  distinct(ICAO, .keep_all = TRUE)

point_to_point_schedule <- full_gen_schedule |>
  mutate(
    dep_time_et = with_tz(dep_time, "America/New_York"),
    arr_time_et = with_tz(arr_time, "America/New_York")
  ) |>
  filter(dep_time_et < as.POSIXct("2026-01-08", tz = "America/New_York")) |>
  mutate(
    day_of_week = factor(weekdays(dep_time_et), 
                         levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"), 
                         ordered = TRUE),
    callsign = paste0(airline, str_sub(tail_number, 2, 5)),
    
    departure_time = as_hms(dep_time_et),
    arrival_time = as_hms(arr_time_et),
    
    duration = as.numeric(duration * 60)
  ) |>
  left_join(airport_locs_clean |> rename(dep_latitude = lat, dep_longitude = lon), by = c("departure" = "ICAO")) |>
  left_join(airport_locs_clean |> rename(arr_latitude = lat, arr_longitude = lon), by = c("arrival" = "ICAO")) |>
  dplyr::select(
    airline, callsign, tail_number,
    departure, arrival,
    day_of_week, departure_time, arrival_time, duration,
    dep_latitude, dep_longitude, arr_latitude, arr_longitude
  ) |>
  arrange(day_of_week, departure_time)

saveRDS(point_to_point_schedule, "point_to_point_schedule.rds")
```

```{r}

schedule <- bind_rows(final_formatted, point_to_point_schedule)
```

```{r}

build_flight_frames <- function(schedule, week_start, step_mins = 30) {
  step_by <- paste(step_mins, "mins")
   
  sched <- schedule |>
    mutate(
      day_index = as.integer(day_of_week) - 1L,
      dep_dt = as.POSIXct(week_start, tz = "UTC") + days(day_index) + seconds(as.numeric(departure_time)),
      arr_dt = as.POSIXct(week_start, tz = "UTC") + days(day_index) + 
               days(if_else(as.numeric(arrival_time) < as.numeric(departure_time), 1L, 0L)) + 
               seconds(as.numeric(arrival_time)),
      airline = str_extract(callsign, "^[A-Z]+")
    ) |>
    group_by(day_index, callsign, departure, arrival) |>
    mutate(
      flight_id = cur_group_id(), 
      ) |>
    ungroup() |>
    filter(
      is.finite(dep_latitude), is.finite(dep_longitude),
      is.finite(arr_latitude), is.finite(arr_longitude),
      arr_dt > dep_dt
    )
 
  # 2. Calculate Great Circle positions based on Time Fractions
  frames <- sched |>
    transmute(flight_id, callsign, airline, dep_dt, arr_dt, 
              dep_lon = dep_longitude, dep_lat = dep_latitude, 
              arr_lon = arr_longitude, arr_lat = arr_latitude) |>
    mutate(
      path_data = pmap(list(dep_lon, dep_lat, arr_lon, arr_lat, dep_dt, arr_dt), 
                       function(d_lon, d_lat, a_lon, a_lat, d_t, a_t) {
                         
        # Generate the time sequence
        t_seq <- seq(
          from = lubridate::floor_date(d_t, unit = step_by), 
          to   = lubridate::ceiling_date(a_t, unit = step_by), 
          by   = step_by
        )
        
        # Filter to actual flight duration and ensure at least start/end exist
        t_seq <- t_seq[t_seq >= d_t & t_seq <= a_t]
        if (length(t_seq) < 2) {
           t_seq <- c(d_t, a_t)
        }
        
        # --- ROBUST CALCULATION METHOD ---
        
        # 1. Calculate total Great Circle distance and initial bearing
        #    (distHaversine returns meters, bearing returns degrees)
        p1 <- c(d_lon, d_lat)
        p2 <- c(a_lon, a_lat)
        total_dist <- distHaversine(p1, p2)
        init_bear  <- bearing(p1, p2)
        
        # 2. Calculate fraction of flight complete at each time step
        total_duration <- as.numeric(difftime(a_t, d_t, units = "secs"))
        
        # Handle zero-duration edge case
        if (total_duration <= 0) {
          return(tibble(time = t_seq, lon = d_lon, lat = d_lat))
        }
        
        time_elapsed <- as.numeric(difftime(t_seq, d_t, units = "secs"))
        fracs <- time_elapsed / total_duration
        
        # 3. Project position using destPoint (Start + Bearing + Distance)
        #    This works on vectors, so we pass the vector of distances
        distances <- total_dist * fracs
        coords <- destPoint(p1, init_bear, distances)
        
        tibble(
          time = t_seq,
          lon = coords[,1],
          lat = coords[,2]
        )
      })
    ) |>
    unnest(path_data) |>
    dplyr::select(flight_id, callsign, airline, dep_dt, arr_dt, time, lat, lon)
 
  return(frames)
}

# --- build a trailing "wake" that disappears after landing ---
build_trail_layers <- function(frames,
                               step_mins = 30,
                               trail_mins = 120) {
  step_seconds <- step_mins * 60
  k <- ceiling(trail_mins / step_mins)

  offsets <- tibble(off_s = (0:k) * step_seconds)

  # propagate each observed point forward into future frames within the trail window
  trail_pts <- frames |>
    inner_join(offsets, by = character()) |>
    mutate(frame_time = time + seconds(off_s)) |>
    filter(frame_time >= dep_dt, frame_time <= arr_dt) |>
    dplyr::select(flight_id, callsign, airline, dep_dt, arr_dt, frame_time, time, lat, lon) |>
    arrange(flight_id, frame_time, time)

  # turn trail points into segments for smoother "line" trails
  trail_segs <- trail_pts |>
    group_by(flight_id, frame_time) |>
    arrange(time, .by_group = TRUE) |>
    mutate(
      lat2 = lead(lat),
      lon2 = lead(lon),
      time2 = lead(time)
    ) |>
    filter(!is.na(lat2), !is.na(lon2)) |>
    # optional fade: newer segments are more opaque
    mutate(
      age_s = as.numeric(difftime(frame_time, time2, units = "secs")),
      alpha = 0.25 * (1 - pmin(1, age_s / (trail_mins * 60)))
    ) |>
    ungroup()

  # current plane position at each frame_time (latest time within that frame)
  current_pts <- trail_pts |>
    group_by(flight_id, frame_time) |>
    slice_max(time, n = 1, with_ties = FALSE) |>
    ungroup()

  list(trail_segs = trail_segs, current_pts = current_pts)
}

frames <- build_flight_frames(
  schedule,
  week_start = as.Date("2025-10-13"),
  step_mins = 30
)

step_mins <- 30

grid <- seq(
  from = as.POSIXct("2025-10-13 00:00:00", tz = "UTC"),
  to   = as.POSIXct("2025-10-20 00:00:00", tz = "UTC"),
  by   = paste(step_mins, "mins")
)

frame_lookup <- tibble(
  frame_time = grid,
  frame_id   = seq_along(grid)
)

layers <- build_trail_layers(frames, step_mins = step_mins, trail_mins = 120)

layers$trail_segs <- layers$trail_segs |>
  left_join(frame_lookup, by = "frame_time")

layers$current_pts <- layers$current_pts |>
  left_join(frame_lookup, by = "frame_time")

layers$trail_segs <- layers$trail_segs |>
  filter(
    !is.na(frame_id),
    is.finite(lon), is.finite(lat),
    is.finite(lon2), is.finite(lat2)
  )

layers$current_pts <- layers$current_pts |>
  filter(!is.na(frame_id))
```

```{r}

us_map <- map_data("state")

p <- ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    linewidth = 0.1,
    color = "grey40",
    fill  = "grey10"
  ) +
  geom_segment(
    data = layers$trail_segs,
    aes(
      x = lon, y = lat,
      xend = lon2, yend = lat2,
      group = interaction(flight_id, frame_time),
      alpha = alpha
    ),
    color = "white",
    linewidth = 0.35,
    lineend = "round"
  ) +
  scale_alpha_identity() +
  geom_point(
    data = layers$current_pts,
    aes(x = lon, y = lat, group = flight_id),
    color = "white",
    alpha = 0.9,
    size = 0.8
  ) +
  coord_quickmap(xlim = c(-125, -66), ylim = c(24, 50)) +
  theme_minimal(base_size = 12) +
  theme(
    panel.background  = element_rect(fill = "white", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    text       = element_text(color = "black"),
    axis.text  = element_text(color = "black"),
    axis.title = element_text(color = "black")
  ) +
  labs(
    title = "{format(
      as.POSIXct('2025-12-22 04:00:00', tz = 'America/New_York') +
        lubridate::minutes(30) * (as.integer(current_frame) - 10),
      '%a %H:%M'
    )} ET"
  ) +
  transition_manual(frame_id) +
  ease_aes("linear")

animate(
  p,
  nframes = length(grid),
  fps = 15,
  width = 1000,
  height = 600,
  renderer = gifski_renderer()
)
```

```{r}

library(truncnorm) # For realistic delay distributions

# -------------------------------------------------------------------------
# 1. PARAMETER DEFINITIONS
# -------------------------------------------------------------------------
# These align with your "Start an airline" requirements

sim_params <- list(
  # Turnaround Times (minutes)
  min_turn_hub = 45,      # Time needed at a hub
  min_turn_spoke = 30,    # Time needed at a spoke/P2P
  
  # Block Time Padding (buffer built into the schedule)
  block_buffer_pct = 0.10, # 10% padding added to ideal flight time
  
  # Disruption Probabilities (The "Random, non-location specific disruption")
  prob_minor_delay = 0.20, # 20% chance of a minor issue (1-15 mins)
  prob_major_delay = 0.05, # 5% chance of a major issue (15-120 mins)
  
  # Airport Capacity Limits (Simplified for "Start Small")
  # Max departures per hour allowed before congestion delays kick in
  capacity_tiers = list(
    "large_airport" = 60,
    "medium_airport" = 30
  )
)

# -------------------------------------------------------------------------
# 2. HELPER FUNCTIONS: DISRUPTION GENERATORS
# -------------------------------------------------------------------------

# Generate random operational delay (crew late, maintenance, baggage)
generate_exogenous_delay <- function(n, params) {
  # Base: Most flights leave on time (0 delay)
  delays <- rep(0, n)
  
  # Minor delays: Truncated Normal distribution
  is_minor <- runif(n) < params$prob_minor_delay
  delays[is_minor] <- rtruncnorm(sum(is_minor), a=1, b=30, mean=10, sd=5)
  
  # Major delays: Gamma distribution (long tail)
  is_major <- runif(n) < params$prob_major_delay
  delays[is_major] <- rgamma(sum(is_major), shape=2, scale=15) + 15
  
  return(round(delays))
}

# Generate en-route variability (Headwinds/Tailwinds)
generate_block_variability <- function(n, scheduled_duration) {
  # Normal distribution centered on 0, sd related to flight length
  noise <- rnorm(n, mean = 0, sd = sqrt(scheduled_duration) * 0.5)
  return(round(noise))
}
```

```{r}

run_network_simulation <- function(input_schedule, params) {
  
  # 1. Prepare the Data
  # We need to ensure the schedule is sorted by Aircraft and Time
  sim_data <- input_schedule |>
    arrange(tail_number, departure_time) |>
    group_by(tail_number) |>
    mutate(
      leg_sequence = row_number(),
      
      # Calculate Scheduled durations (Plan)
      sched_dep_min = as.numeric(departure_time) / 60,
      sched_arr_min = as.numeric(arrival_time) / 60,
      sched_block   = sched_arr_min - sched_dep_min,
      
      # Identify Turn Type (Hub vs Spoke) for minimum turn time logic
      required_turn = if_else(arrival %in% c("KATL", "KORD", "KDEN", "KDFW", "KCLT"), 
                              params$min_turn_hub, 
                              params$min_turn_spoke)
    ) |>
    ungroup()
  
  # Initialize storage for the "Actual" results
  n_flights <- nrow(sim_data)
  
  # Pre-calculate exogenous delays (Vectorized for speed)
  sim_data$exo_delay_dep <- generate_exogenous_delay(n_flights, params)
  sim_data$block_noise   <- generate_block_variability(n_flights, sim_data$sched_block)
  
  # 2. The Simulation Loop (Iterative by Tail Number)
  # We use a loop because Leg N depends on the outcome of Leg N-1
  
  # Split by tail number for faster processing
  flights_by_tail <- split(sim_data, sim_data$tail_number)
  
  results_list <- lapply(flights_by_tail, function(plane_schedule) {
    
    # Initialize columns
    plane_schedule$act_dep_min <- NA
    plane_schedule$act_arr_min <- NA
    plane_schedule$prop_delay  <- 0 # Propagated delay
    
    for (i in 1:nrow(plane_schedule)) {
      
      # Step A: Determine Earliest Possible Departure
      if (i == 1) {
        # First flight of the day: Only affected by exogenous delay (e.g., crew late)
        ready_time <- 0
      } else {
        # Subsequent flights: Ready when previous flight arrived + Turn Time
        ready_time <- plane_schedule$act_arr_min[i-1] + plane_schedule$required_turn[i-1]
      }
      
      # Step B: Calculate Actual Departure
      # Plane leaves at Scheduled Time OR Ready Time (whichever is later) + Exogenous Delay
      sched_time <- plane_schedule$sched_dep_min[i]
      
      # Calculate Propagated Delay (The "Knock-on" effect)
      # If Ready Time > Scheduled Time, that difference is propagated delay
      prop_delay <- max(0, ready_time - sched_time)
      
      plane_schedule$prop_delay[i] <- prop_delay
      
      # Final Departure Time
      plane_schedule$act_dep_min[i] <- sched_time + prop_delay + plane_schedule$exo_delay_dep[i]
      
      # Step C: Calculate Actual Arrival
      # Actual Dep + Scheduled Block + Padding Buffer + Enroute Noise
      # Note: If they fly faster (negative noise), they can recover some delay
      actual_block <- plane_schedule$sched_block[i] + plane_schedule$block_noise[i]
      
      plane_schedule$act_arr_min[i] <- plane_schedule$act_dep_min[i] + actual_block
    }
    
    return(plane_schedule)
  })
  
  # Recombine
  final_results <- bind_rows(results_list)
  
  return(final_results)
}
```

```{r}

# Filter for Hub Carriers from your generated schedule
hub_schedule <- schedule |>
  filter(airline %in% c("UAL", "DAL", "AAL")) 

# Run Simulation
hub_results <- run_network_simulation(hub_schedule, sim_params)

# Calculate Metrics
hub_metrics <- hub_results |>
  summarise(
    Scenario = "Hub-and-Spoke",
    Total_Flights = n(),
    Avg_Arr_Delay = mean(act_arr_min - sched_arr_min),
    Propagated_Delay_Share = sum(prop_delay) / sum(act_arr_min - sched_arr_min)
  )

print(hub_metrics)
```

```{r}

# Filter for P2P Carriers
p2p_schedule <- schedule |>
  filter(airline %in% c("SWA", "NKS", "JBU", "FFT"))

# Run Simulation
p2p_results <- run_network_simulation(p2p_schedule, sim_params)

# Calculate Metrics
p2p_metrics <- p2p_results |>
  summarise(
    Scenario = "Point-to-Point",
    Total_Flights = n(),
    Avg_Arr_Delay = mean(act_arr_min - sched_arr_min),
    Propagated_Delay_Share = sum(prop_delay) / sum(act_arr_min - sched_arr_min)
  )

print(p2p_metrics)
```

```{r}

visualize_propagation <- function(sim_output, target_tail) {
  
  data <- sim_output |>
    filter(tail_number == target_tail) |>
    arrange(departure_time) |>
    mutate(
      leg_id = paste(departure, ">", arrival),
      delay_status = case_when(
        prop_delay > 15 ~ "Propagated Delay",
        exo_delay_dep > 15 ~ "New Disruption",
        TRUE ~ "On Time / Recovered"
      )
    )
  
  ggplot(data) +
    # Plan
    geom_segment(aes(x = sched_dep_min/60, xend = sched_arr_min/60, 
                     y = leg_sequence, yend = leg_sequence), 
                 color = "grey", size = 2, alpha = 0.5) +
    # Actual
    geom_segment(aes(x = act_dep_min/60, xend = act_arr_min/60, 
                     y = leg_sequence, yend = leg_sequence, color = delay_status), 
                 size = 1.5) +
    scale_y_reverse(breaks = data$leg_sequence, labels = data$leg_id) +
    labs(
      title = paste("Delay Propagation for Aircraft:", target_tail),
      x = "Hour of Day (Local)",
      y = "Flight Leg",
      color = "Delay Source"
    ) +
    theme_minimal()
}

# Example usage (pick a tail number from your results)
visualize_propagation(hub_results, "N00001DAL")
```
